---
title: "<span style='color: red; font-size: 30px;'>k Nearest Neighbors (kNN) - Fundamentals</span>"
title-block-banner-color: "red"
author: "Dr. Aydede"
subtitle: "MBAN - Winter 2024"
format: html
toc: true
toc-float: true
number-sections: true
theme: united
highlight: tango
self_contained: true
---

# Distances

In k-Nearest Neighbors (kNN), the concept of distance is fundamental for determining the nearest neighbors of a given data point. Several distance metrics can be used, each with its characteristics and suitability for different types of data. The main distance metrics commonly used in kNN are:

## Euclidean Distance:

The Euclidean distance is the straight-line distance between the two points. It is the most commonly used distance metric in kNN. The Euclidean distance between two points $\mathbf{x}=\left(x_{1}, x_{2}\right)$ and $\mathbf{y}=\left(y_{1}, y_{2}\right)$ is given by:

$$
d\left(\mathbf{x}, \mathbf{y}\right)=\sqrt{\left(x_{1}-x_{2}\right)^2+\left(y_{1}-y_{2}\right)^2}
$$

```{r}
# Euclidean Distance
EDistance <- function(x, y){
  n <- length(x)
  dd <- matrix(0, n, n)
  
  for (i in 1:n) {
    for (j in 1:n) {
      dd[i, j] <- sqrt((x[i] - x[j])^2 + (y[i] - y[j])^2)
    }
  }
  return(dd)
}

# Example
x1 <- c(2, 2.1, 4, 4.3)
x2 <- c(3, 3.3, 5, 5.1)
EDistance(x1, x2)
```

## Manhattan Distance:

The Manhattan distance is the sum of the absolute differences between the two points. The Manhattan distance between two points $\mathbf{x}=\left(x_{1}, x_{2}\right)$ and $\mathbf{y}=\left(y_{1}, y_{2}\right)$ is given by:

$$
d\left(\mathbf{x}, \mathbf{y}\right)=\left|x_{1}-x_{2}\right|+\left|y_{1}-y_{2}\right|
$$

```{r}
# Manhattan Distance
MDistance <- function(x, y){
  n <- length(x)
  dd <- matrix(0, n, n)
  
  for (i in 1:n) {
    for (j in 1:n) {
      dd[i, j] <- abs(x[i] - x[j]) + abs(y[i] - y[j])
    }
  }
  return(dd)
}
```

## Chebyshev Distance:

The Chebyshev distance is the maximum absolute difference between the two points. The Chebyshev distance between two points $\mathbf{x}=\left(x_{1}, x_{2}\right)$ and $\mathbf{y}=\left(y_{1}, y_{2}\right)$ is given by:

$$
d\left(\mathbf{x}, \mathbf{y}\right)=\max \left(\left|x_{1}-x_{2}\right|,\left|y_{1}-y_{2}\right|\right)
$$

```{r}
# Chebyshev Distance
CDistance <- function(x, y){
  n <- length(x)
  dd <- matrix(0, n, n)
  
  for (i in 1:n) {
    for (j in 1:n) {
      dd[i, j] <- max(abs(x[i] - x[j]), abs(y[i] - y[j]))
    }
  }
  return(dd)
}
```

# Example: finding NN with k = 2

```{r}
x1 <- c(2, 2.1, 4, 4.3, 1.5, 3.2, 6, 5.5, 7, 2.5, 4.5, 3.8, 6.2, 5, 7.3)
x2 <- c(3, 3.3, 5, 5.1, 2, 4.2, 6.5, 5.8, 7.2, 3.1, 4.7, 3.6, 6.1, 5.2, 7.6)

distanceMatrix <- EDistance(x1, x2)
print(distanceMatrix)
```

Given that distance matrix, we can find the nearest neighbors of each point. For example, the nearest neighbors of the first point are the second and fifth points. The nearest neighbors of the second point are the first and fifth points, and so on. Let's write a function to find the nearest neighbors of each point.

```{r}
findNN <- function(distanceMatrix, k) {
  n <- nrow(distanceMatrix)
  nearestNeighbors <- matrix(NA, n, k)
  
  for (i in 1:n) {
    # Sort distances
    sortedDistances <- sort(distanceMatrix[i, ], index.return = TRUE)
    
    # Get the indices of the nearest neighbors (first two elements after the first)
    nnIndices <- sortedDistances$ix[2:(k + 1)]
    nearestNeighbors[i,] <- nnIndices
  }
  return(nearestNeighbors)
}

distanceMatrix <- EDistance(x1, x2)
nearestNeighbors <- findNN(distanceMatrix, 2)
print(nearestNeighbors)
```

# More realistics example

Our job is to predict the labels of the new data points (blue points) based on the labels of the training data (red and black points). We will develop a very simple kNN algorithm to predict the labels of the new data points.

```{r, fig.width=6, fig.height=6}
n = 25

# Our data
set.seed(12)
x1 <- runif(n)  # 15 random values for x1
x2 <- runif(n)  # 15 random values for x2
y <- sample(c(0, 1), n, replace = TRUE)  

# New data - with unknown labels
set.seed(222)
x1_new <- runif(5)  
x2_new <- runif(5)  

# Plot the data
colors <- ifelse(y == 1, "red", "black")
plot(x1, x2, col = colors, xlab = "x1", ylab = "x2", lwd = 2)
points(x1_new, x2_new, col = "blue", pch = 4, lwd = 2)
```

Let's try with k = 5. For each new data point, we will find the 5 nearest neighbors and assign the most common label to the new data point.

```{r}
k <- 5  
new_labels <- c()

for (i in 1:length(x1_new)) {
  # Calculate distances for the i-th new point to all old points
  distances <- sqrt((x1_new[i] - x1)^2 + (x2_new[i] - x2)^2)
  
  # Find the k nearest points
  nearest <- order(distances)[1:k]
  
  # Find the labels of the k nearest points
  nearest_labels <- y[nearest]
  
  # Find the most common label
  new_labels[i] <- names(which.max(table(nearest_labels)))
}

print(new_labels)

# Plot the new data with the new labels
colors <- ifelse(y == 1, "red", "black")
plot(x1, x2, col = colors, xlab = "x1", ylab = "x2", lwd = 2)
colors_new <- ifelse(new_labels == 1, "red", "black")
points(x1_new, x2_new, col = colors_new, pch = 4, lwd = 2)
```

# `mnist_27` dataset

```{r}
#loading the data
library(tidyverse)
library(dslabs)
library(parallel)

data("mnist_27")
str(mnist_27)

#Scale the data.  
mdata <- mnist_27$train
mdata$x_1 <- scale(mdata$x_1)
mdata$x_2 <- scale(mdata$x_2)
```

Split `mdata` into training and validation sets. For now we split it 75% and 25%

```{r}
set.seed(1)
ind <- sample(nrow(mdata), nrow(mdata) * 0.75, replace = FALSE)
train <- mdata[ind, ]
val <- mdata[-ind, ]

# Plot the training data and validation data together
colors <- ifelse(train$y == 7, "red", "black")
plot(train$x_1, train$x_2, col = colors, xlab = "x1", ylab = "x2", lwd = 1)
points(val$x_1, val$x_2, col = "green", pch = 4, lwd = 0.75)
```

We want to train a model that identifies the green points (validation set) very well. Let's have a function:

```{r}
# Function to determine the label for a validation set
find_label <- function(new_point, k){
  distances <- sqrt((new_point[1] - train[, 2])^2 + (new_point[2] - train[, 3])^2)
  nearest <- order(distances)[1:k]
  nearest_labels <- train$y[nearest]
  return(names(which.max(table(nearest_labels))))
}
```

Now let's apply this function to the validation set and calculate the accuracy. Here accuracy is defined as the proportion of correct labels in the validation set. This is a correct way to measure accuracy when the classes are balanced (i.e. there are roughly the same number of 0's and 1's).

```{r}
# Pick your k
new_data <- val[, 2:3]
new_labels <- apply(new_data, 1, function(x) find_label(x, k = 30))
acc <- mean(new_labels == val$y)
acc
```

## Single-run grid search for k

```{r, cache=TRUE}
# Same function as before, but now we pass k as an argument
find_label <- function(new_point, k){
  distances <- sqrt((new_point[1] - train[, 2])^2 + (new_point[2] - train[, 3])^2)
  nearest <- order(distances)[1:k]
  nearest_labels <- train$y[nearest]
  return(names(which.max(table(nearest_labels))))
}

# to compute accuracy for a grid of k's
acc <- c()
grid <- seq(5, 100, by = 2)

for (i in 1:length(grid)) {
  k <- grid[i]
  set.seed(1)
  ind <- unique(sample(nrow(mdata), nrow(mdata), replace = TRUE))
  train <- mdata[ind, ]
  val <- mdata[-ind, ]
  new_data <- val[, 2:3]
  new_labels <- apply(new_data, 1, function(x) find_label(x, k = k))
  acc[i] <- mean(new_labels == val$y)
}

plot(grid, acc, type = "l", xlab = "k", ylab = "Accuracy")
```

We can show that there is another way to do that with `sapply`:

```{r}
# Same function as before
find_label <- function(new_point, k){
  distances <- sqrt((new_point[1] - train[, 2])^2 + (new_point[2] - train[, 3])^2)
  nearest <- order(distances)[1:k]
  nearest_labels <- train$y[nearest]
  return(names(which.max(table(nearest_labels))))
}

# Function to compute accuracy with splits
compute_accuracy <- function(k) {
  set.seed(1)
  ind <- unique(sample(nrow(mdata), nrow(mdata), replace = TRUE))
  train <- mdata[ind, ]
  val <- mdata[-ind, ]
  new_data <- val[, 2:3]
  new_labels <- apply(new_data, 1, function(x) find_label(x, k = k))
  return(mean(new_labels == val$y))
}

# Compute accuracy for different values of k using `sapply`
grid <- seq(5, 100, by = 2)
acc <- sapply(grid, compute_accuracy)
plot(grid, acc, type = "l", xlab = "k", ylab = "Accuracy")
```

## Multi-run grid search (training the model) for k

```{r}
find_label <- function(new_point, k){
  distances <- sqrt((new_point[1] - train[, 2])^2 + (new_point[2] - train[, 3])^2)
  nearest <- order(distances)[1:k]
  nearest_labels <- train$y[nearest]
  return(names(which.max(table(nearest_labels))))
}

macc <- c()
u95 <- c()
l95 <- c()
grid <- seq(5, 100, by = 2)

for (k in 1:length(grid)) {

  acc <- c()
  
  for (i in 1:30) {
    set.seed(i)
    ind <- unique(sample(nrow(mdata), nrow(mdata), replace = TRUE))
    train <- mdata[ind, ]
    val <- mdata[-ind, ]
    new_data <- val[, 2:3]
    new_labels <- apply(new_data, 1, function(x) find_label(x, k = grid[k]))
    acc[i] <- mean(new_labels == val$y)
  }  
  macc[k] <- mean(acc)
}

plot(grid, macc, type = "l", xlab = "k", ylab = "Accuracy")
```

## Grid search with parallel computing (`parLapply`)


```{r}
library(parallel)

compute_accuracy <- function(params) {
  k <- params$k
  seed <- params$seed
  
  set.seed(seed)  # Set seed for reproducibility
  ind <- unique(sample(nrow(mdata), nrow(mdata), replace = TRUE))
  train <- mdata[ind, ]
  val <- mdata[-ind, ]
  new_data <- val[, 2:3]
  
  find_label <- function(new_point, k, train_data) {
    distances <- sqrt((new_point[1] - train_data[, 2])^2 + (new_point[2] - train_data[, 3])^2)
    nearest <- order(distances)[1:k]
    nearest_labels <- train_data$y[nearest]
    return(names(which.max(table(nearest_labels))))
  }
  
  new_labels <- apply(new_data, 1, function(x) find_label(x, k, train))
  return(mean(new_labels == val$y))
}

# Parallel implementation for Windows
macc_parallel <- c()
grid <- seq(5, 120, by = 1)
no_cores <- detectCores() - 1
cl <- makeCluster(no_cores)

# Export the required data and functions to all clusters
clusterExport(cl, c("mdata", "compute_accuracy", "grid"))

for (k in 1:length(grid)) {
  # Create parameter list with consistent seeds
  params_list <- lapply(1:30, function(i) list(k = grid[k], seed = i))
  
  # Run parallel computation using parLapply instead of mclapply
  acc <- unlist(parLapply(cl, params_list, compute_accuracy))
  macc_parallel[k] <- mean(acc)
}

stopCluster(cl)

# Plot results
plot(grid, macc_parallel, type = "o", xlab = "k", ylab = "Accuracy", 
     main = "Parallel Implementation (Windows)", col = "blue", lwd = 1.5)
abline(v = grid[which.max(macc_parallel)], lty = 2, col = "red")
```

# Ready to use packages

There are several packages in R for kNN applications: `knn()` from the `class` package and `knn3()` in the caret package. Since kNN use distances, we should scale the numerical variables first to make their magnitudes on the same scale. The `knn()` function does not do this for us, so we have to do it ourselves. The `knn3()` function in the caret package does this for us. Note also that while `knn3()` handles factor variables itself, which is an internal process, a good one, `knn()` could not do that and requires all features to be numeric.

## `knn3` (from `caret`) - single vs parallel core application

We can use parallel computing to speed it up. The `doParallel` package must be loaded for this to work. The `foreach` package is also useful for parallel computing. The `foreach()` function is similar to the `lapply()` function, but it uses multiple cores.


```{r}
library(caret)
library(foreach)
library(doParallel)

# Define consistent parameters
grid <- seq(5, 120, by = 1)
n_iterations <- 30

# Function to compute accuracy that will be used in both implementations
compute_knn_accuracy <- function(k, seed) {
  set.seed(seed)
  ind <- unique(sample(nrow(mdata), nrow(mdata), replace = TRUE))
  train <- mdata[ind, ]
  val <- mdata[-ind, ]
  
  model <- knn3(y ~ ., data = train, k = k)
  yhat <- predict(model, val, type = "class")
  return(mean(yhat == val$y))
}

# Sequential Implementation
macc_sequential <- numeric(length(grid))
for (i in seq_along(grid)) {
  acc <- numeric(n_iterations)
  for (j in 1:n_iterations) {
    acc[j] <- compute_knn_accuracy(grid[i], j)
  }
  macc_sequential[i] <- mean(acc)
}

# Parallel Implementation
no_cores <- detectCores() - 1
cl <- makeCluster(no_cores)
registerDoParallel(cl)

# Create parameter combinations
param_grid <- expand.grid(k = grid, seed = 1:n_iterations)
macc_parallel <- numeric(length(grid))

# Make sure each cluster has the required packages - suppress output
invisible(clusterEvalQ(cl, {
  suppressPackageStartupMessages(library(caret))
  NULL
}))

# Export necessary objects to the cluster
clusterExport(cl, c("mdata", "compute_knn_accuracy"))

# Use foreach for parallel processing
for (i in seq_along(grid)) {
  k_params <- param_grid[param_grid$k == grid[i], ]
  
  acc <- foreach(j = 1:n_iterations, .combine = 'c') %dopar% {
    compute_knn_accuracy(k_params$k[j], k_params$seed[j])
  }
  
  macc_parallel[i] <- mean(acc)
}

stopCluster(cl)

# Compare results visually
par(mfrow = c(1, 2))
plot(grid, macc_sequential, type = "o", 
     xlab = "k", ylab = "Accuracy",
     main = "Sequential Implementation",
     col = "blue", lwd = 1.5)
abline(v = grid[which.max(macc_sequential)], lty = 2, col = "red")
text(x = 70, y = 0.82, 
     labels = paste("k =", grid[which.max(macc_sequential)], 
                   "\nAccuracy =", round(max(macc_sequential), 5)),
     pos = 4, col = "red")

plot(grid, macc_parallel, type = "o", 
     xlab = "k", ylab = "Accuracy",
     main = "Parallel Implementation",
     col = "blue", lwd = 1.5)
abline(v = grid[which.max(macc_parallel)], lty = 2, col = "red")
text(x = 70, y = 0.82, 
     labels = paste("k =", grid[which.max(macc_parallel)], 
                   "\nAccuracy =", round(max(macc_parallel), 5)),
     pos = 4, col = "red")

# Print summary statistics
cat("\nSequential Implementation - Optimal k:", grid[which.max(macc_sequential)],
    "\nParallel Implementation - Optimal k:", grid[which.max(macc_parallel)])
```

```{r}
# registerDoSEQ() is a function in R that registers the sequential (non-parallel) backend for the foreach package. (sets stuff back to normal)
registerDoSEQ()
```



## `train` by `caret` package

We can use `train`, the automated function for training in `caret`. By default, the cross validation is performed by taking 25 bootstrap samples comprised of 25% of the observations. Moreover, the default is to try k (5,7,9). We can to expand it:

```{r, cache=TRUE, warning=FALSE, message=FALSE}
library(caret)
library(foreach)
library(doParallel)

# Define consistent parameters across both approaches
grid <- seq(5, 120, by = 1)
n_iterations <- 25  # Match train()'s default
sample_prop <- 0.75 # Proportion for training data

# Modified implementation to match train()'s validation approach
compute_knn_accuracy <- function(k, seed) {
    set.seed(seed)
    # Sample indices without replacement to match train()'s approach
    train_indices <- sample(nrow(mdata), size = floor(nrow(mdata) * sample_prop))
    train_data <- mdata[train_indices, ]
    val_data <- mdata[-train_indices, ]
    
    model <- knn3(y ~ ., data = train_data, k = k)
    yhat <- predict(model, val_data, type = "class")
    return(mean(yhat == val_data$y))
}

# Sequential implementation with aligned validation
macc_sequential <- numeric(length(grid))
for (i in seq_along(grid)) {
    acc <- numeric(n_iterations)
    for (j in 1:n_iterations) {
        acc[j] <- compute_knn_accuracy(grid[i], j)
    }
    macc_sequential[i] <- mean(acc)
}

# Train implementation with matched parameters
set.seed(123)  # For reproducibility
train_control <- trainControl(
    method = "cv",
    number = n_iterations,
    p = sample_prop
)

train_knn <- train(
    y ~ ., 
    method = "knn",
    data = mdata,
    tuneGrid = data.frame(k = grid),
    trControl = train_control
)

# Compare results visually
par(mfrow = c(1, 2))
plot(grid, macc_sequential, type = "o", 
     xlab = "k", ylab = "Accuracy",
     main = "Manual Implementation",
     col = "blue", lwd = 1.5)
abline(v = grid[which.max(macc_sequential)], lty = 2, col = "red")

plot(train_knn$results$k, train_knn$results$Accuracy, 
     type = "o", xlab = "k", ylab = "Accuracy",
     main = "train() Implementation",
     col = "blue", lwd = 1.5)
abline(v = train_knn$bestTune$k, lty = 2, col = "red")

# Print optimal k values
cat("\nManual Implementation Optimal k:", grid[which.max(macc_sequential)],
    "\ntrain() Implementation Optimal k:", train_knn$bestTune$k)
```

Key alignments in this implementation:

- Same sampling proportion (75% for training)
- Same number of iterations (25)
- Consistent cross-validation approach using sampling without replacement
- Same grid of k values
- Same random seed handling

This modification should provide more comparable results between the two methods. The main remaining difference would be in how train() handles internal computational details, but the fundamental validation structure is now aligned.

## `class` package

```{r, cache=TRUE}
# Training with kNN
library(class)

macc <- c()
grid <- seq(9, 120, by = 1)

for (k in 1:length(grid)) {
  for (i in 1:30) {
    set.seed(i)
    ind <- unique(sample(nrow(mdata), nrow(mdata), replace = TRUE))
    train <- mdata[ind, ]
    val <- mdata[-ind, ]
    knnmodel <- knn(train[, 2:3], val[, 2:3], train$y, k = grid[k])
    acc[i] <- mean(knnmodel == val$y)
  }
  macc[k] <- mean(acc)
}

# Plotting the accuracy
plot(grid, macc, type = "l", xlab = "k", ylab = "Accuracy")
```
