---
title: "<span style='color: red; font-size: 30px;'>GBM vs XGBoost: Applied Comparison</span>"
title-block-banner-color: "red"
author: "Dr. Aydede"
subtitle: "MBAN 5560 - Winter 2026"
format:
  html:
    embed-resources: true
    code-background: true
toc: true
toc-float: true
number-sections: true
theme: united
highlight: tango
---

In the previous notebooks, we covered the theory behind GBM and XGBoost separately. Now let's put them to work on the **same real-world dataset** and compare them head-to-head — first for **regression**, then for **classification**.

We'll use the **Ames Housing** dataset (2,930 home sales in Ames, Iowa) with 80 features describing every aspect of a house. This is the same dataset you've seen in earlier assignments, but now we'll use it to compare two boosting algorithms side by side.

# Data Preparation

Both GBM and XGBoost need the same data, but in different formats. Let's load and prepare everything once, then feed it to both algorithms.

## Loading and Cleaning

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(gbm)
library(xgboost)
library(caret)
library(ROCR)

# Load the Ames Housing data
ames <- read.csv("../Week4/Assignment/AmesHousing.csv")

# Clean column names: replace spaces and special characters
names(ames) <- make.names(names(ames), unique = TRUE)

dim(ames)
```

```{r}
# Quick look at the target variable
summary(ames$SalePrice)
hist(ames$SalePrice, breaks = 50, col = "steelblue", main = "Sale Price Distribution",
     xlab = "Sale Price ($)")
```

The distribution is right-skewed — common for price data. We'll use `log(SalePrice)` for regression to make the distribution more symmetric.

## Feature Selection and Cleaning

The full dataset has 82 columns. Many have missing values or low information. Let's select a practical subset of features that cover different aspects of a house:

```{r}
# Select useful features + target
features <- c(
  "Gr.Liv.Area", "Overall.Qual", "Overall.Cond", "Year.Built",
  "Total.Bsmt.SF", "X1st.Flr.SF", "Full.Bath", "Half.Bath",
  "Bedroom.AbvGr", "Garage.Cars", "Garage.Area", "Lot.Area",
  "Fireplaces", "TotRms.AbvGrd", "Year.Remod.Add",
  "MS.Zoning", "Neighborhood", "Bldg.Type", "House.Style",
  "Central.Air", "Foundation", "Exter.Qual"
)

df <- ames[, c(features, "SalePrice")]

# Remove rows with any NA
df <- na.omit(df)

# Convert character columns to factors (required by gbm)
df <- df %>% mutate(across(where(is.character), as.factor))

dim(df)
```

```{r}
# Log-transform the target for regression
df$logSalePrice <- log(df$SalePrice)
```

## Train/Test Split

We use a fixed 80/20 split so both GBM and XGBoost are evaluated on the **exact same** test set:

```{r}
set.seed(42)
train_idx <- sample(1:nrow(df), size = 0.8 * nrow(df))
train_df <- df[train_idx, ]
test_df  <- df[-train_idx, ]

cat("Training:", nrow(train_df), "rows\n")
cat("Test:    ", nrow(test_df), "rows\n")
```

## Preparing XGBoost Format

GBM can work directly with data frames and formulas. XGBoost needs a numeric matrix. We prepare both:

```{r}
# One-hot encode for XGBoost using caret::dummyVars()
dv <- dummyVars(~ . - SalePrice - logSalePrice, data = train_df)

X_train <- predict(dv, newdata = train_df)
X_test  <- predict(dv, newdata = test_df)

cat("Feature matrix:", ncol(X_train), "columns after one-hot encoding\n")
```

```{r}
# Create DMatrix objects for XGBoost
# Regression targets (log scale)
dtrain_reg <- xgb.DMatrix(data = X_train, label = train_df$logSalePrice)
dtest_reg  <- xgb.DMatrix(data = X_test,  label = test_df$logSalePrice)
```

---

# Regression — Predicting Sale Price

**Goal**: Predict `log(SalePrice)` using housing features.
**Metric**: RMSE on the test set (on log scale).

## GBM Regression

We tune GBM using its built-in 10-fold cross-validation to find the optimal number of trees:

```{r, cache=TRUE, message=FALSE, warning=FALSE}
t_gbm <- system.time({
  gbm_reg <- gbm(
    formula = logSalePrice ~ . - SalePrice,
    distribution = "gaussian",
    data = train_df,
    n.trees = 2000,
    interaction.depth = 4,
    shrinkage = 0.05,
    n.minobsinnode = 10,
    bag.fraction = 0.8,
    cv.folds = 10,
    verbose = FALSE
  )
})

# Best iteration from CV
best_iter_gbm <- gbm.perf(gbm_reg, method = "cv", plot.it = FALSE)
cat("GBM best iteration:", best_iter_gbm, "\n")
cat("GBM CV RMSE:       ", round(sqrt(min(gbm_reg$cv.error)), 4), "\n")
cat("GBM training time: ", round(t_gbm["elapsed"], 1), "seconds\n")
```

### GBM Test Predictions

```{r}
# Predict on test set
yhat_gbm <- predict(gbm_reg, newdata = test_df, n.trees = best_iter_gbm)

# RMSE on log scale
rmse_gbm <- sqrt(mean((test_df$logSalePrice - yhat_gbm)^2))
cat("GBM test RMSE (log scale):", round(rmse_gbm, 4), "\n")
```

## XGBoost Regression

We use `xgb.train()` with early stopping on a validation set:

```{r, cache=TRUE, message=FALSE, warning=FALSE}
evals_reg <- list(train = dtrain_reg, test = dtest_reg)

params_reg <- list(
  max_depth = 4,
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8,
  objective = "reg:squarederror",
  eval_metric = "rmse"
)

t_xgb <- system.time({
  xgb_reg <- xgb.train(
    params = params_reg,
    data = dtrain_reg,
    nrounds = 2000,
    evals = evals_reg,
    early_stopping_rounds = 20,
    verbose = 0
  )
})

best_iter_xgb <- as.numeric(xgb.attributes(xgb_reg)$best_iteration)
best_rmse_xgb <- as.numeric(xgb.attributes(xgb_reg)$best_score)
cat("XGBoost best iteration:", best_iter_xgb, "\n")
cat("XGBoost best val RMSE: ", round(best_rmse_xgb, 4), "\n")
cat("XGBoost training time: ", round(t_xgb["elapsed"], 1), "seconds\n")
```

### XGBoost Test Predictions

```{r}
yhat_xgb <- predict(xgb_reg, dtest_reg)

rmse_xgb <- sqrt(mean((test_df$logSalePrice - yhat_xgb)^2))
cat("XGBoost test RMSE (log scale):", round(rmse_xgb, 4), "\n")
```

## Regression Head-to-Head

```{r}
# Summary comparison
reg_comparison <- data.frame(
  Model = c("GBM", "XGBoost"),
  Best_Iteration = c(best_iter_gbm, best_iter_xgb),
  Test_RMSE = round(c(rmse_gbm, rmse_xgb), 4),
  Training_Time_sec = round(c(t_gbm["elapsed"], t_xgb["elapsed"]), 1)
)
reg_comparison
```

### Variable Importance Comparison

```{r, fig.height=5, fig.width=10}
par(mfrow = c(1, 2))

# GBM importance
gbm_imp <- summary.gbm(gbm_reg, plotit = FALSE)
barplot(head(gbm_imp$rel.inf, 10), names.arg = head(gbm_imp$var, 10),
        las = 2, col = "steelblue", main = "GBM: Top 10 Features",
        ylab = "Relative Influence", cex.names = 0.7)

# XGBoost importance
xgb_imp <- xgb.importance(model = xgb_reg)
barplot(head(xgb_imp$Gain, 10), names.arg = head(xgb_imp$Feature, 10),
        las = 2, col = "coral", main = "XGBoost: Top 10 Features",
        ylab = "Gain", cex.names = 0.7)
```

Notice something odd? In GBM, `Neighborhood` is the **2nd most important variable**. In XGBoost's top 10, it's **completely missing**. Did XGBoost decide neighborhoods don't matter? No — this is an artifact of **one-hot encoding**.

GBM handles categorical variables natively: it treats `Neighborhood` as one variable and reports its total importance. XGBoost requires a numeric matrix, so we one-hot encoded `Neighborhood` into ~25 dummy columns (`Neighborhood.Northridge`, `Neighborhood.Old_Town`, etc.). The importance of `Neighborhood` gets **fragmented** across all these dummies — each one contributes a small slice, so none individually ranks in the top 10.


::: {.callout-note}
## Regression results

Both models produce similar test RMSE — they use the same gradient boosting framework underneath. The key differences are:

**Speed.** XGBoost is typically faster, especially on larger datasets, because it parallelizes the split search within each tree.

**Tuning workflow: CV vs. validation set.** GBM uses `cv.folds` for internal k-fold cross-validation — the approach you already know well. XGBoost takes a different approach: you pass a held-out **validation set** via the `evals` argument and set `early_stopping_rounds`. During training, XGBoost evaluates the metric on this validation set after every round. If the metric doesn't improve for `early_stopping_rounds` consecutive rounds, it stops. This is conceptually similar to CV (both estimate out-of-sample error to pick the best iteration), but using a single validation set is faster than running k separate folds. XGBoost also supports `xgb.cv()` for k-fold CV when you want it (we use it in the grid search section below).

**Feature importance and one-hot encoding.** This is the most important difference to understand. GBM handles categorical variables natively — `Neighborhood` is one variable, and its total influence is reported as a single number. XGBoost requires a numeric matrix, so categorical variables must be one-hot encoded first. This means `Neighborhood` becomes ~25 separate dummy columns, and its importance gets **fragmented** across all of them. As the plots above show, `Neighborhood` is the 2nd most important feature in GBM but completely missing from XGBoost's raw top 10. This doesn't mean XGBoost thinks neighborhoods don't matter — it means the importance is spread thinly across many dummies, so none individually ranks high. **Takeaway:** when comparing feature importance across models, keep in mind that one-hot encoding can make important categorical variables "disappear" from the top of XGBoost's list. To get a fair comparison, you would need to aggregate the dummy-level importance back to the original variable level.
:::

---

# Classification — Predicting High vs. Low Price

Now we convert the same data into a **classification problem**: is the sale price above or below the median?

## Creating the Binary Target

```{r}
# Create binary target: 1 = above median, 0 = below median
median_price <- median(train_df$SalePrice)
cat("Median sale price (from training data): $", format(median_price, big.mark = ","), "\n")

train_df$price_high <- ifelse(train_df$SalePrice > median_price, 1, 0)
test_df$price_high  <- ifelse(test_df$SalePrice > median_price, 1, 0)

cat("\nTraining set class distribution:\n")
table(train_df$price_high)

cat("\nTest set class distribution:\n")
table(test_df$price_high)
```

```{r}
# Update XGBoost DMatrix with classification labels
dtrain_cls <- xgb.DMatrix(data = X_train, label = train_df$price_high)
dtest_cls  <- xgb.DMatrix(data = X_test,  label = test_df$price_high)
```

## GBM Classification

```{r, cache=TRUE, message=FALSE, warning=FALSE}
t_gbm_cls <- system.time({
  gbm_cls <- gbm(
    formula = price_high ~ . - SalePrice - logSalePrice,
    distribution = "bernoulli",
    data = train_df,
    n.trees = 2000,
    interaction.depth = 4,
    shrinkage = 0.05,
    n.minobsinnode = 10,
    bag.fraction = 0.8,
    cv.folds = 10,
    verbose = FALSE
  )
})

best_iter_gbm_cls <- gbm.perf(gbm_cls, method = "cv", plot.it = FALSE)
cat("GBM best iteration:", best_iter_gbm_cls, "\n")
cat("GBM training time: ", round(t_gbm_cls["elapsed"], 1), "seconds\n")
```

### GBM Classification Results

```{r}
# Predicted probabilities on test set
phat_gbm <- predict(gbm_cls, newdata = test_df, n.trees = best_iter_gbm_cls, type = "response")

# AUC
pred_gbm_cls <- prediction(phat_gbm, test_df$price_high)
auc_gbm <- performance(pred_gbm_cls, "auc")@y.values[[1]]
cat("GBM AUC:", round(auc_gbm, 4), "\n")
```

```{r}
# Confusion matrix
yhat_gbm_cls <- factor(ifelse(phat_gbm > 0.5, 1, 0), levels = c(0, 1))
test_cls_factor <- factor(test_df$price_high, levels = c(0, 1))
confusionMatrix(yhat_gbm_cls, test_cls_factor, positive = "1")
```

## XGBoost Classification

```{r, cache=TRUE, message=FALSE, warning=FALSE}
evals_cls <- list(train = dtrain_cls, test = dtest_cls)

params_cls <- list(
  max_depth = 4,
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8,
  objective = "binary:logistic",
  eval_metric = "auc"
)

t_xgb_cls <- system.time({
  xgb_cls <- xgb.train(
    params = params_cls,
    data = dtrain_cls,
    nrounds = 2000,
    evals = evals_cls,
    early_stopping_rounds = 20,
    verbose = 0
  )
})

best_iter_xgb_cls <- as.numeric(xgb.attributes(xgb_cls)$best_iteration)
best_auc_xgb <- as.numeric(xgb.attributes(xgb_cls)$best_score)
cat("XGBoost best iteration:", best_iter_xgb_cls, "\n")
cat("XGBoost best val AUC:  ", round(best_auc_xgb, 4), "\n")
cat("XGBoost training time: ", round(t_xgb_cls["elapsed"], 1), "seconds\n")
```

### XGBoost Classification Results

```{r}
# Predicted probabilities on test set
phat_xgb <- predict(xgb_cls, dtest_cls)

# AUC
pred_xgb_cls <- prediction(phat_xgb, test_df$price_high)
auc_xgb <- performance(pred_xgb_cls, "auc")@y.values[[1]]
cat("XGBoost AUC:", round(auc_xgb, 4), "\n")
```

```{r}
# Confusion matrix
yhat_xgb_cls <- factor(ifelse(phat_xgb > 0.5, 1, 0), levels = c(0, 1))
confusionMatrix(yhat_xgb_cls, test_cls_factor, positive = "1")
```

## Classification Head-to-Head

```{r}
cls_comparison <- data.frame(
  Model = c("GBM", "XGBoost"),
  Best_Iteration = c(best_iter_gbm_cls, best_iter_xgb_cls),
  Test_AUC = round(c(auc_gbm, auc_xgb), 4),
  Training_Time_sec = round(c(t_gbm_cls["elapsed"], t_xgb_cls["elapsed"]), 1)
)
cls_comparison
```

### ROC Curves

```{r, fig.width=7, fig.height=5}
# ROC curves for both models
roc_gbm <- performance(pred_gbm_cls, "tpr", "fpr")
roc_xgb <- performance(pred_xgb_cls, "tpr", "fpr")

plot(roc_gbm, col = "steelblue", lwd = 2, main = "ROC Curves: GBM vs XGBoost")
plot(roc_xgb, col = "coral", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey")
legend("bottomright",
       legend = c(paste0("GBM (AUC = ", round(auc_gbm, 3), ")"),
                  paste0("XGBoost (AUC = ", round(auc_xgb, 3), ")")),
       col = c("steelblue", "coral"), lwd = 2)
```

---

# Can We Do Better? Hyperparameter Grid Search

In the previous sections, we used a single set of "reasonable" hyperparameters for both models. But how do we know those are the best? In practice, you should **systematically search** over combinations of hyperparameters to find the best configuration. This is called **grid search**.

We'll demonstrate this for the **regression** task (predicting log SalePrice). The same approach works for classification — just change the loss function and evaluation metric.

## GBM Grid Search

GBM's built-in `cv.folds` makes grid search straightforward: loop over parameter combinations, use internal CV to find the best iteration for each, and record the CV error.

```{r, cache=TRUE, message=FALSE, warning=FALSE}
# Define the grid
gbm_grid <- expand.grid(
  shrinkage = c(0.01, 0.05, 0.1),
  interaction.depth = c(3, 4, 5),
  n.minobsinnode = c(5, 10),
  bag.fraction = c(0.7, 0.85),
  best_trees = 0,
  cv_rmse = 0
)

cat("Total GBM combinations to try:", nrow(gbm_grid), "\n")
```

```{r, cache=TRUE, message=FALSE, warning=FALSE}
t_gbm_grid <- system.time({
  for (i in 1:nrow(gbm_grid)) {
    
    fit <- gbm(
      formula = logSalePrice ~ . - SalePrice,
      distribution = "gaussian",
      data = train_df,
      n.trees = 2000,
      interaction.depth = gbm_grid$interaction.depth[i],
      shrinkage = gbm_grid$shrinkage[i],
      n.minobsinnode = gbm_grid$n.minobsinnode[i],
      bag.fraction = gbm_grid$bag.fraction[i],
      cv.folds = 5,
      verbose = FALSE
    )
    
    gbm_grid$best_trees[i] <- which.min(fit$cv.error)
    gbm_grid$cv_rmse[i] <- sqrt(min(fit$cv.error))
  }
})

cat("GBM grid search time:", round(t_gbm_grid["elapsed"], 1), "seconds\n\n")

# Show top 10
gbm_grid %>%
  arrange(cv_rmse) %>%
  head(10)
```

```{r}
# Best GBM configuration
best_gbm <- gbm_grid %>% arrange(cv_rmse) %>% slice(1)
cat("Best GBM hyperparameters:\n")
cat("  shrinkage:        ", best_gbm$shrinkage, "\n")
cat("  interaction.depth:", best_gbm$interaction.depth, "\n")
cat("  n.minobsinnode:   ", best_gbm$n.minobsinnode, "\n")
cat("  bag.fraction:     ", best_gbm$bag.fraction, "\n")
cat("  best_trees:       ", best_gbm$best_trees, "\n")
cat("  CV RMSE:          ", round(best_gbm$cv_rmse, 4), "\n")
```

### Tuned GBM: Test Performance

```{r, cache=TRUE, message=FALSE, warning=FALSE}
# Retrain with best hyperparameters (no CV needed — just the final model)
gbm_tuned <- gbm(
  formula = logSalePrice ~ . - SalePrice,
  distribution = "gaussian",
  data = train_df,
  n.trees = best_gbm$best_trees,
  interaction.depth = best_gbm$interaction.depth,
  shrinkage = best_gbm$shrinkage,
  n.minobsinnode = best_gbm$n.minobsinnode,
  bag.fraction = best_gbm$bag.fraction,
  verbose = FALSE
)

yhat_gbm_tuned <- predict(gbm_tuned, newdata = test_df, n.trees = best_gbm$best_trees)
rmse_gbm_tuned <- sqrt(mean((test_df$logSalePrice - yhat_gbm_tuned)^2))
cat("Tuned GBM test RMSE:", round(rmse_gbm_tuned, 4), "\n")
cat("Baseline GBM test RMSE:", round(rmse_gbm, 4), "\n")
```

## XGBoost Grid Search

In the regression section above, we used `xgb.train()` with a held-out validation set (`evals` + `early_stopping_rounds`) to pick the best iteration. That's fast but relies on a single train/validation split. For grid search, we switch to `xgb.cv()`, which performs **k-fold cross-validation internally** — just like GBM's `cv.folds`. It splits the training data into k folds, trains k separate models, averages the error across folds, and supports `early_stopping_rounds` so each configuration stops early when it plateaus. This gives a more reliable estimate of each hyperparameter combination's quality, which is exactly what we need when comparing many configurations.

```{r, cache=TRUE, message=FALSE, warning=FALSE}
# Define the grid
xgb_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1),
  max_depth = c(3, 4, 5),
  subsample = c(0.7, 0.85),
  colsample_bytree = c(0.7, 0.85),
  best_nrounds = 0,
  cv_rmse = 0
)

cat("Total XGBoost combinations to try:", nrow(xgb_grid), "\n")
```

```{r, cache=TRUE, message=FALSE, warning=FALSE}
t_xgb_grid <- system.time({
  for (i in 1:nrow(xgb_grid)) {
    
    params_i <- list(
      eta = xgb_grid$eta[i],
      max_depth = xgb_grid$max_depth[i],
      subsample = xgb_grid$subsample[i],
      colsample_bytree = xgb_grid$colsample_bytree[i],
      objective = "reg:squarederror",
      eval_metric = "rmse"
    )
    
    cv_result <- xgb.cv(
      params = params_i,
      data = dtrain_reg,
      nrounds = 2000,
      nfold = 5,
      early_stopping_rounds = 20,
      verbose = 0
    )
    
    # Extract best result from evaluation log
    eval_log <- as.data.frame(cv_result$evaluation_log)
    test_col <- grep("test.*rmse.*mean", names(eval_log), value = TRUE)[1]
    best_idx <- which.min(eval_log[[test_col]])
    
    xgb_grid$best_nrounds[i] <- best_idx
    xgb_grid$cv_rmse[i] <- eval_log[[test_col]][best_idx]
  }
})

cat("XGBoost grid search time:", round(t_xgb_grid["elapsed"], 1), "seconds\n\n")

# Show top 10
xgb_grid %>%
  arrange(cv_rmse) %>%
  head(10)
```

```{r}
# Best XGBoost configuration
best_xgb <- xgb_grid %>% arrange(cv_rmse) %>% slice(1)
cat("Best XGBoost hyperparameters:\n")
cat("  eta:              ", best_xgb$eta, "\n")
cat("  max_depth:        ", best_xgb$max_depth, "\n")
cat("  subsample:        ", best_xgb$subsample, "\n")
cat("  colsample_bytree: ", best_xgb$colsample_bytree, "\n")
cat("  best_nrounds:     ", best_xgb$best_nrounds, "\n")
cat("  CV RMSE:          ", round(best_xgb$cv_rmse, 4), "\n")
```

### Tuned XGBoost: Test Performance

```{r, cache=TRUE, message=FALSE, warning=FALSE}
# Retrain with best hyperparameters on full training data
params_best <- list(
  eta = best_xgb$eta,
  max_depth = best_xgb$max_depth,
  subsample = best_xgb$subsample,
  colsample_bytree = best_xgb$colsample_bytree,
  objective = "reg:squarederror",
  eval_metric = "rmse"
)

xgb_tuned <- xgb.train(
  params = params_best,
  data = dtrain_reg,
  nrounds = best_xgb$best_nrounds,
  verbose = 0
)

yhat_xgb_tuned <- predict(xgb_tuned, dtest_reg)
rmse_xgb_tuned <- sqrt(mean((test_df$logSalePrice - yhat_xgb_tuned)^2))
cat("Tuned XGBoost test RMSE:", round(rmse_xgb_tuned, 4), "\n")
cat("Baseline XGBoost test RMSE:", round(rmse_xgb, 4), "\n")
```

## Grid Search: Did Tuning Help?

```{r}
tuning_comparison <- data.frame(
  Model = c("GBM (baseline)", "GBM (tuned)", "XGBoost (baseline)", "XGBoost (tuned)"),
  Test_RMSE = round(c(rmse_gbm, rmse_gbm_tuned, rmse_xgb, rmse_xgb_tuned), 4),
  Grid_Search_Time = c("—",
                       paste0(round(t_gbm_grid["elapsed"], 1), "s"),
                       "—",
                       paste0(round(t_xgb_grid["elapsed"], 1), "s"))
)
tuning_comparison
```

::: {.callout-note}
## Grid search takeaways

1. **Tuning usually helps** — even small improvements in RMSE matter when multiplied across many predictions.
2. **GBM grid search is slower** because each `cv.folds` run trains 5 models × 2000 trees for every combination. XGBoost's `xgb.cv()` with early stopping is faster because it stops training early when the error plateaus.
3. **You don't need to search everything at once**: Start with a coarse grid (few values, wide range), find the best region, then zoom in with a finer grid. This is called **two-stage grid search**.
4. **The grid above is small** (36 combinations). In practice you might try more values, but even this small grid shows the workflow clearly.
:::

---

# Summary

```{r}
# Final comparison table
final <- data.frame(
  Task = c("Regression", "Regression", "Classification", "Classification"),
  Model = c("GBM", "XGBoost", "GBM", "XGBoost"),
  Metric = c("RMSE", "RMSE", "AUC", "AUC"),
  Value = round(c(rmse_gbm, rmse_xgb, auc_gbm, auc_xgb), 4),
  Training_Time = round(c(t_gbm["elapsed"], t_xgb["elapsed"],
                          t_gbm_cls["elapsed"], t_xgb_cls["elapsed"]), 1)
)
final
```

::: {.callout-note}
## Key Takeaways

1. **Predictive accuracy is similar**: Both GBM and XGBoost use the same gradient boosting framework, so on the same data with similar hyperparameters, they produce comparable results.

2. **XGBoost is faster**: The speed advantage becomes more pronounced with larger datasets. On this ~2,300-row dataset, the difference may be modest; on 100K+ rows it would be dramatic.

3. **Different workflows**:
   - GBM: Pass a formula + data frame → use `cv.folds` for tuning → `gbm.perf()` for best iteration
   - XGBoost: Convert to matrix → create `xgb.DMatrix` → use `evals` + `early_stopping_rounds` → `xgb.attributes()` for best iteration

4. **XGBoost has more tuning knobs**: `lambda`, `alpha`, and `colsample_bytree` give you additional regularization options beyond what GBM offers.

5. **Data prep is the real work**: Both algorithms required the same data cleaning, feature selection, and train/test split. The modeling part is fast — the prep is where the effort goes.
:::
