---
title: "<span style='color: red; font-size: 50px; font-weight: bold;'>Neural Networks: The Idea</span>"
subtitle: "From Polynomial Regression to Adaptive Basis Functions"
author: "Dr. Aydede"
date: today
format: 
  html:
    embed-resources: true
    code-background: true
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    fig-width: 9
    fig-height: 6
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(patchwork)
```

Neural networks (NN) are a class of models inspired by the human brain. **They are, however, parametric models**, and they are used for supervised learning. In this note, we build the intuition for how neural networks work by starting from something familiar — polynomial regression — and showing how neural networks generalize it.

# The Idea

## Polynomial Regression

Let's start with a predictive model with a single input (covariate). The simplest model could be a linear model:

$$
y \approx \alpha+\beta x
$$

Since this model could be quite restrictive, we can have a more flexible one through polynomial regression:

$$
y \approx \alpha+\beta_1 x+\beta_2 x^2+\beta_3 x^3+\ldots=\alpha+\sum_{m=1}^M \beta_m x^m
$$

The polynomial regression is based on **fixed components**, or bases: $x, x^2, x^3, \ldots, x^M$. These are predetermined transformations of the input — we don't learn them from data.

Let us consider a realistic (simulated) sample:

```{r sim-data}
n <- 200
set.seed(1)
x <- sort(runif(n))
y <- sin(12 * (x + 0.2)) / (x + 0.2) + rnorm(n) / 2
df <- data.frame(y, x)

ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "grey60", alpha = 0.7, size = 2) +
  labs(title = "Simulated Data",
       subtitle = "True function: sin(12(x + 0.2)) / (x + 0.2) + noise",
       x = "x", y = "y") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

We can fit a polynomial regression with M = 3:

```{r poly-fit}
ols <- lm(y ~ x + I(x^2) + I(x^3))

ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "grey60", alpha = 0.7, size = 2) +
  geom_line(aes(y = predict(ols)), color = "blue", linewidth = 1.5) +
  labs(title = "Polynomial Regression: M = 3",
       subtitle = expression(y == alpha + beta[1]*x + beta[2]*x^2 + beta[3]*x^3),
       x = "x", y = "y") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

Now, we can think of the line as a weighted sum of **fixed components**: $\alpha + \beta_1 x + \beta_2 x^2 + \beta_3 x^3$.

```{r poly-components}
# Decompose into individual components
first <- ols$coefficients[2] * x
second <- ols$coefficients[3] * x^2
third <- ols$coefficients[4] * x^3

components_df <- data.frame(
  x = rep(x, 3),
  value = c(first, second, third),
  Component = rep(c("b1 * x", "b2 * x^2", "b3 * x^3"), each = n)
)

ggplot(components_df, aes(x = x, y = value, color = Component)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("b1 * x" = "pink", "b2 * x^2" = "orange", "b3 * x^3" = "green")) +
  facet_wrap(~Component, scales = "free_y") +
  labs(title = "Fixed Components of the Polynomial",
       subtitle = "Each component is a predetermined transformation of x",
       x = "x", y = "Component value") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "none")
```

And their sum:

```{r poly-sum}
yhat <- ols$coefficients[1] + first + second + third

ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "grey60", alpha = 0.7, size = 2) +
  geom_line(aes(y = yhat), color = "red", linewidth = 1.5) +
  labs(title = expression(y == alpha + beta[1]*x + beta[2]*x^2 + beta[3]*x^3),
       subtitle = "Sum of Fixed Components",
       x = "x", y = "y") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

## Polynomial to Neural Network

The artificial neural net replaces these fixed components with **adjustable ones (adaptive bases)**:

$$
f\left(\alpha_1+\delta_1 x\right), \quad f\left(\alpha_2+\delta_2 x\right), \quad \ldots, \quad f\left(\alpha_M+\delta_M x\right)
$$ 
  
where $f(.)$ is an **activation function**. We can see the first simple ANN as nonlinear functions of linear combinations:

$$
\begin{gathered}
y \approx \alpha+\beta_1 f\left(\alpha_1+\delta_1 x\right)+\beta_2 f\left(\alpha_2+\delta_2 x\right)+\beta_3 f\left(\alpha_3+\delta_3 x\right)+\ldots \\
=\alpha+\sum_{m=1}^M \beta_m f\left(\alpha_m+\delta_m x\right)
\end{gathered}
$$

The activation function $f(.)$ introduces nonlinearity. Common choices include:

```{r activation-table, echo=FALSE}
act_df <- data.frame(
  Function = c("Sigmoid (Logistic)", "Hyperbolic Tangent (tanh)", "ReLU"),
  Formula = c("f(x) = 1 / (1 + e⁻ˣ)", "f(x) = (eˣ - e⁻ˣ) / (eˣ + e⁻ˣ)", "f(x) = max(0, x)"),
  Range = c("(0, 1)", "(-1, 1)", "[0, ∞)"),
  `Key_Property` = c(
    "Smooth S-shaped curve, outputs probabilities",
    "Zero-centered, stronger gradients than sigmoid",
    "Simple, efficient, dominant in deep learning"
  )
)

kable(act_df,
      caption = "Common Activation Functions",
      col.names = c("Function", "Formula", "Output Range", "Key Property")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

```{r activation-viz, echo=FALSE}
x_act <- seq(-5, 5, length.out = 200)

act_data <- data.frame(
  x = rep(x_act, 3),
  y = c(
    1 / (1 + exp(-x_act)),           # Sigmoid
    tanh(x_act),                      # Tanh
    pmax(0, x_act)                    # ReLU
  ),
  Function = rep(c("Sigmoid", "Tanh", "ReLU"), each = 200)
)

ggplot(act_data, aes(x = x, y = y, color = Function)) +
  geom_line(linewidth = 1.3) +
  facet_wrap(~Function, scales = "free_y") +
  scale_color_manual(values = c("Sigmoid" = "#E41A1C", "Tanh" = "#377EB8", "ReLU" = "#4DAF4A")) +
  labs(title = "Common Activation Functions",
       x = "Input (z)", y = "Output f(z)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "none")
```

The crucial difference from polynomial regression: **the parameters inside the activation function ($\alpha_m$ and $\delta_m$) are learned from the data**, not predetermined. This means the network can **adapt its basis functions** to best capture the patterns in the data. Adjustable components enable capturing complex models with fewer components (smaller M).

## Demonstration: ANN with Arbitrary Parameters

Let's replace those fixed components $x, x^2, x^3$ in our polynomial regression with $f(\alpha_1 + \delta_1 x)$, $f(\alpha_2 + \delta_2 x)$, $f(\alpha_3 + \delta_3 x)$.

The following code demonstrates the ability of a simple artificial neural network (ANN) with **arbitrary** (hand-picked) parameters to capture the underlying signal relative to a third-degree polynomial regression model. For each node, the code:

1. Computes the weighted input: $z_m = \alpha_m + \delta_m \cdot x$
2. Applies the sigmoid activation: $f(z_m) = \frac{1}{1 + e^{-z_m}}$
3. Multiplies by the output weight: $\beta_m \cdot f(z_m)$

The final output is: $\hat{y} = \text{intercept} + \beta_1 f(z_1) + \beta_2 f(z_2) + \beta_3 f(z_3)$

```{r ann-arbitrary}
a <- c(1.5, 9, 3)
b <- c(-20, -14, -8)
beta <- c(15, 25, -40)
int <- 3

ann <- function(a, b, beta, int) {
  # 1st sigmoid node
  z1 <- a[1] + b[1] * x
  f1 <- 1 / (1 + exp(-z1))
  
  # 2nd sigmoid node
  z2 <- a[2] + b[2] * x
  f2 <- 1 / (1 + exp(-z2))
  
  # 3rd sigmoid node
  z3 <- a[3] + b[3] * x
  f3 <- 1 / (1 + exp(-z3))
  
  yhat <- int + beta[1] * f1 + beta[2] * f2 + beta[3] * f3
  return(yhat)
}

yhat_ann <- ann(a, b, beta, int)

ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "grey60", alpha = 0.7, size = 2) +
  geom_line(aes(y = yhat_ann), color = "red", linewidth = 1.5) +
  labs(title = "ANN with Arbitrary Parameters: M = 3 Nodes",
       subtitle = "Hand-picked parameters — not yet optimized",
       x = "x", y = "y") +
  ylim(-5, 15) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

Now let's see what happens when we **learn** the parameters from data using `neuralnet`:

```{r ann-trained}
library(neuralnet)
set.seed(2)
nn <- neuralnet(y ~ x, data = df, hidden = 3, threshold = 0.05)
yhat_nn <- compute(nn, data.frame(x))$net.result

ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "grey60", alpha = 0.7, size = 2) +
  geom_line(aes(y = as.vector(yhat_nn)), color = "red", linewidth = 1.5) +
  labs(title = "Neural Network with Learned Parameters: M = 3 Nodes",
       subtitle = "Parameters optimized via backpropagation",
       x = "x", y = "y") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

Why did the neural network perform better than polynomial regression? Because **adjustable components adapt to the data**, while fixed polynomial bases are rigid. Let's look deeper at the learned weight structure.

## Understanding the Learned Weights

The model is:

$$
y \approx \alpha + \sum_{m=1}^{3} \beta_m \cdot f(\alpha_m + \delta_m x)
$$

$$
= \alpha + \beta_1 f(\alpha_1 + \delta_1 x) + \beta_2 f(\alpha_2 + \delta_2 x) + \beta_3 f(\alpha_3 + \delta_3 x)
$$

Here are the learned weights:

```{r nn-weights}
nn$weights
```

```{r nn-plot}
plot(nn, rep = "best")
```

- **Error**: The final error of the trained network — how well the model has learned the underlying pattern
- **Steps**: The number of iterations (weight updates) the algorithm took to converge

## Visualizing the Adaptive Basis Functions

We used sigmoid (logistic) activation functions. Using the learned weights:

- Node 1: $f(x) = \frac{1}{1 + e^{-(1.263 - 18.959 \cdot x)}}$
- Node 2: $f(x) = \frac{1}{1 + e^{-(6.600 - 12.247 \cdot x)}}$
- Node 3: $f(x) = \frac{1}{1 + e^{-(2.505 - 5.701 \cdot x)}}$

Each node creates a different "building block" — a shifted and scaled sigmoid:

```{r nn-components}
X <- cbind(1, x)

# Compute activation for each node
n1 <- nn$weights[[1]][[1]][, 1]
f1 <- nn$act.fct(X %*% n1)

n2 <- nn$weights[[1]][[1]][, 2]
f2 <- nn$act.fct(X %*% n2)

n3 <- nn$weights[[1]][[1]][, 3]
f3 <- nn$act.fct(X %*% n3)

# Visualize the adaptive basis functions
basis_df <- data.frame(
  x = rep(x, 3),
  value = c(as.vector(f1), as.vector(f2), as.vector(f3)),
  Node = rep(c("Node 1: f(a1 + d1*x)", "Node 2: f(a2 + d2*x)", "Node 3: f(a3 + d3*x)"), each = n)
)

ggplot(basis_df, aes(x = x, y = value, color = Node)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("pink", "orange", "green")) +
  facet_wrap(~Node) +
  labs(title = "Adaptive Basis Functions (Learned from Data)",
       subtitle = "Compare with the fixed polynomial components — these shapes are learned!",
       x = "x", y = "Activation output") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "none")
```

Now we go from these nodes to the output. Each node's activation is multiplied by its output weight ($\beta_m$), and the results are summed with a bias (intercept):

$$
\hat{y} = \underbrace{2.408}_{\text{bias}} + \underbrace{13.032}_{\beta_1} \cdot f_1 + \underbrace{19.924}_{\beta_2} \cdot f_2 + \underbrace{(-32.173)}_{\beta_3} \cdot f_3
$$

```{r nn-output}
# From nodes to output
f12 <- f1 * nn$weights[[1]][[2]][2]
f22 <- f2 * nn$weights[[1]][[2]][3]
f32 <- f3 * nn$weights[[1]][[2]][4]

# Final prediction
yhat_manual <- nn$weights[[1]][[2]][1] + f12 + f22 + f32

ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "grey60", alpha = 0.7, size = 2) +
  geom_line(aes(y = as.vector(yhat_manual)), color = "red", linewidth = 1.5) +
  labs(title = "ANN Prediction: Manually Reconstructed from Weights",
       subtitle = "ŷ = bias + β₁·f₁ + β₂·f₂ + β₃·f₃",
       x = "x", y = "y") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

# Neural Networks with Multiple Inputs

With a set of covariates $X = (1, x_1, x_2, \ldots, x_k)$, the model becomes:

$$
y \approx \alpha + \sum_{m=1}^{M} \beta_m f\left(\alpha_m + \delta_{m1} x_1 + \delta_{m2} x_2 + \cdots + \delta_{mk} x_k\right)
$$

Each hidden node now takes a **weighted combination of all input features**, passes it through the activation function, and the output layer combines all the node outputs.

## Example: Boston Housing Data

Let's apply a neural network to a real dataset:

```{r boston-data}
library(MASS)
data("Boston")

# Quick summary
kable(data.frame(
  Feature = names(Boston),
  Type = sapply(Boston, class),
  Mean = round(sapply(Boston, mean), 2),
  SD = round(sapply(Boston, sd), 2),
  Min = round(sapply(Boston, min), 2),
  Max = round(sapply(Boston, max), 2)
), caption = "Boston Housing Dataset Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12)
```

```{r boston-prep}
# Check for missing values
cat("Missing values:", sum(is.na(Boston)), "\n")

# Normalize all features (neural networks are sensitive to scale)
normalizedBoston <- as.data.frame(scale(Boston))

# Train-test split
set.seed(123)
ind <- sample(1:nrow(normalizedBoston), size = 0.8 * nrow(normalizedBoston))
train_data <- normalizedBoston[ind, ]
test_data <- normalizedBoston[-ind, ]

# Separate targets
train_targets <- train_data$medv
test_targets <- test_data$medv
train_data$medv <- NULL
test_data$medv <- NULL
```

## Training a Neural Network

```{r boston-nn}
library(neuralnet)

# Define the formula
formula <- paste("train_targets ~", paste(names(train_data), collapse = " + "))

# Train the neural network with 5 hidden neurons
set.seed(123)
nn1 <- neuralnet(formula, 
                 data = cbind(train_targets, train_data), 
                 hidden = c(5), 
                 linear.output = TRUE)
```

Here, `hidden = c(5)` defines a neural network with **one hidden layer containing 5 neurons**. `linear.output = TRUE` is used because this is a regression problem (continuous output).

```{r boston-nn-plot}
plot(nn1, rep = "best")
```

## Prediction Performance

```{r boston-pred}
mse_test1 <- mean((test_targets - predict(nn1, test_data))^2)
cat(sprintf("Test MSE (5 hidden neurons): %.4f\n", mse_test1))
```

Is it good? Let's try a different architecture:

```{r boston-nn2}
# Try with 4 hidden neurons and more training steps
set.seed(123)
nn2 <- neuralnet(formula, 
                 data = cbind(train_targets, train_data), 
                 hidden = c(4), 
                 linear.output = TRUE, 
                 stepmax = 1e5, 
                 learningrate = 0.01)
```

```{r boston-nn2-plot}
plot(nn2, rep = "best")
```

```{r boston-nn2-pred}
mse_test2 <- mean((test_targets - predict(nn2, test_data))^2)

# Compare the two models
results_df <- data.frame(
  Model = c("NN (5 hidden neurons)", "NN (4 hidden neurons)"),
  Test_MSE = c(mse_test1, mse_test2)
)

kable(results_df,
      caption = "Comparing Neural Network Architectures",
      col.names = c("Model", "Test MSE"),
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Convergence Issues

If you see the warning `Algorithm did not converge`, it means the optimization didn't find a stable solution within the allowed steps. This can happen for several reasons:

```{r convergence-table, echo=FALSE}
conv_df <- data.frame(
  Cause = c(
    "Insufficient training steps",
    "Learning rate too high/low",
    "Too complex architecture",
    "Data scaling issues"
  ),
  Solution = c(
    "Increase stepmax parameter",
    "Adjust learningrate (try 0.01, 0.001)",
    "Reduce hidden layers/neurons",
    "Ensure all features are scaled (we did this)"
  ),
  Intuition = c(
    "Give the algorithm more time to converge",
    "Too high → overshoots; too low → too slow",
    "Simpler models are easier to optimize",
    "Features on different scales create elongated loss surfaces"
  )
)

kable(conv_df,
      caption = "Common Convergence Issues and Solutions",
      col.names = c("Cause", "Solution", "Intuition")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

# Understanding `neuralnet` Arguments

## Core Arguments

```{r core-args-table, echo=FALSE}
core_args <- data.frame(
  Argument = c("formula", "data", "hidden", "threshold", "stepmax", "rep"),
  Description = c(
    "Model specification: response ~ predictors",
    "Data frame containing all variables",
    "Vector specifying neurons per hidden layer. c(5,3) = two layers with 5 and 3 neurons",
    "Stopping criterion: stops when all partial derivatives of error are below this value",
    "Maximum number of training steps (iterations)",
    "Number of training repetitions with different random starting weights"
  ),
  Typical_Values = c(
    "—",
    "—",
    "c(5), c(10,5), c(8,4,2)",
    "0.01 (default), 0.05 for faster convergence",
    "1e5 to 1e6",
    "1-5 (more reps = better chance of good solution)"
  )
)

kable(core_args,
      caption = "Core Arguments of neuralnet()",
      col.names = c("Argument", "Description", "Typical Values")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE, color = "darkblue")
```

## Learning Rate Arguments

```{r lr-args-table, echo=FALSE}
lr_args <- data.frame(
  Argument = c("startweights", "learningrate.limit", "learningrate.factor", "learningrate"),
  Description = c(
    "Initial weights (NULL = random initialization)",
    "Min/max bounds for learning rate (for backprop)",
    "Factors for increasing/decreasing learning rate (for rprop)",
    "Fixed learning rate (for backprop algorithm)"
  ),
  Default = c("NULL", "c(1e-10, 0.1)", "list(minus=0.5, plus=1.2)", "NULL")
)

kable(lr_args,
      caption = "Learning Rate Arguments",
      col.names = c("Argument", "Description", "Default")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE, color = "darkblue")
```

## Algorithm and Output Control

```{r algo-args-table, echo=FALSE}
algo_args <- data.frame(
  Argument = c("algorithm", "err.fct", "act.fct", "linear.output"),
  Description = c(
    "Training algorithm: 'rprop+', 'rprop-', 'backprop', or 'slr'",
    "Error function: 'sse' (sum of squared errors) or 'ce' (cross-entropy)",
    "Activation function: 'logistic' or 'tanh'",
    "If TRUE, output neuron uses linear activation (for regression)"
  ),
  When_To_Use = c(
    "rprop+ (default) is usually best; backprop for custom learning rates",
    "sse for regression, ce for classification",
    "logistic (default) or tanh — both are common",
    "TRUE for regression, FALSE for classification"
  )
)

kable(algo_args,
      caption = "Algorithm and Output Control",
      col.names = c("Argument", "Description", "When to Use")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE, color = "darkblue")
```

## Stepmax and Rep in `neuralnet`

In `GradientD.qmd`, we learned three key concepts for gradient descent:

- **Epoch** = one complete pass through the entire training dataset  
- **Batch size** = the number of observations used per parameter update  
- **Iteration** = one parameter update (one epoch may contain many iterations if we use mini-batches)

How does `neuralnet` relate to these? The key fact is:

> **`neuralnet` uses full-batch Rprop by default** — it computes gradients on the **entire training set** at every step. There are no mini-batches.

This means **each "step" in `neuralnet` is equivalent to one epoch**: one complete pass through all the data, followed by one weight update. There is no batch size to set — every step uses all $N$ observations.

With that in mind, `neuralnet` has two important parameters:

### `stepmax`: How long you train (within one run)

Think of training a neural network as hiking downhill to find the lowest valley. Each **step** is one weight update using the entire dataset. `stepmax` says: *"stop walking after this many steps, even if you haven't reached the bottom."*

- `stepmax = 1e5` → you're allowed up to 100,000 steps per training run
- If the algorithm converges (reaches the `threshold`) before `stepmax`, it stops early
- If it hasn't converged by `stepmax`, training halts and you get a convergence warning → increase `stepmax` to give it more time

### `rep`: How many times you start over from scratch

Neural network loss surfaces are **non-convex** — there are many valleys (local minima), not just one. Where you end up depends on **where you start** (the random initial weights). `rep` says: *"start the entire hike over from a different random starting point this many times."*

- `rep = 20` → train the network **20 completely independent times**, each with different random starting weights
- Each of those 20 runs gets its own `stepmax` budget
- Then you can pick the best one (lowest error)

### They are independent (nested loops)

`rep` is the **outer loop** and `stepmax` is the budget **inside each rep**. Here's what happens with `rep = 3, stepmax = 1e5`:

```
Rep 1:  random weights → train up to 100,000 steps → final error = 0.45
Rep 2:  random weights → train up to 100,000 steps → final error = 0.32
Rep 3:  random weights → train up to 100,000 steps → final error = 0.89 (bad start!)
→ Pick the best: Rep 2 (error = 0.32)
```

Total work = 3 runs × up to 100,000 steps each = up to 300,000 steps total.

```{r stepmax-rep-table, echo=FALSE}
concepts <- data.frame(
  neuralnet_Concept = c("Step", "Stepmax", "Rep"),
  What_It_Means = c(
    "One weight update using the entire dataset",
    "Max steps within one training run — the budget per rep",
    "Number of independent training runs, each starting from different random weights"
  ),
  GradientD_Equivalent = c(
    "≈ One epoch (full-batch gradient descent)",
    "≈ Maximum number of epochs per run",
    "No direct equivalent — addresses the local minima problem"
  ),
  Typical_Values = c(
    "—",
    "1e5 to 1e6 (increase if you see convergence warnings)",
    "1–5 (more reps = better chance of finding a good solution)"
  )
)

kable(concepts,
      caption = "Mapping neuralnet Concepts to Gradient Descent Terminology",
      col.names = c("neuralnet Concept", "What It Means", "GradientD Equivalent", "Typical Values")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

# Backpropagation Algorithms

## Resilient Backpropagation (Rprop)

Resilient Backpropagation (Rprop) is the **default algorithm** in `neuralnet`. It overcomes a key difficulty with standard gradient descent: the learning rate problem.

Standard gradient descent updates weights proportional to both the **direction** and **magnitude** of the gradient. But gradient magnitudes can vary wildly across different weights, making a single learning rate problematic.

Rprop's solution is elegant:

1. **Only the sign matters**: Rprop uses only the **sign** of the gradient (positive or negative) to determine the update direction, ignoring the magnitude
2. **Individual step sizes**: Each weight has its own adaptive step size
3. **Adaptive adjustment**: If the gradient keeps the same sign → increase the step size (we're heading in the right direction, go faster). If the gradient changes sign → decrease the step size (we overshot, be more careful)

```{r rprop-table, echo=FALSE}
rprop_df <- data.frame(
  Variant = c("Rprop+", "Rprop-", "iRprop+"),
  Description = c(
    "If gradient changes sign (overshoot), undo the last update and decrease step size",
    "Only adjusts step sizes without reverting the last update",
    "Improved version that handles flat regions of the error surface better"
  ),
  When_To_Use = c(
    "Default in neuralnet — good general choice",
    "Simpler, sometimes faster",
    "When the error surface has many flat regions"
  )
)

kable(rprop_df,
      caption = "Variants of Resilient Backpropagation",
      col.names = c("Variant", "Description", "When to Use")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

The key advantages of Rprop over standard backpropagation:

- **No global learning rate to tune**: Each weight adapts independently
- **Faster convergence**: Adapts quickly to the local curvature of the error surface
- **More robust**: Less sensitive to the scale of inputs and the architecture of the network

---

### References

- Hastie, T., Tibshirani, R., & Friedman, J. (2009). *The Elements of Statistical Learning*. Springer. Chapter 11: Neural Networks.
- Goodfellow, I., Bengio, Y., & Courville, A. (2016). *Deep Learning*. MIT Press. Chapters 6-8.
- Riedmiller, M., & Braun, H. (1993). A direct adaptive method for faster backpropagation learning: The RPROP algorithm. *IEEE International Conference on Neural Networks*.

### R Session Information

```{r session-info}
sessionInfo()
```
