---
title: "<span style='color: red; font-size: 50px; font-weight: bold;'>Deep Learning 101</span>"
subtitle: "From Single-Layer Networks to Deep Neural Networks with Keras"
author: "Dr. Aydede"
date: today
format: 
  html:
    embed-resources: true
    code-background: true
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    fig-width: 9
    fig-height: 6
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(reticulate)
use_virtualenv("/Users/yigitaydede/.virtualenvs/r-tensorflow", required = TRUE)
Sys.unsetenv("TF_USE_LEGACY_KERAS")
```

Deep Learning is an advanced subset of machine learning that mimics the way humans gain certain types of knowledge. At its core, deep learning involves neural networks with many layers, hence the term "deep." This depth allows models to learn complex patterns in large amounts of data, making it a powerful tool for a wide range of applications, from image and speech recognition to natural language processing.

We've already learned **neural networks with a single hidden layer**, which can capture a certain level of complexity in data. These networks consist of **an input layer, one hidden layer, and an output layer**. While simple neural networks are useful for learning linear relationships, their ability to model complex, nonlinear relationships is limited.

**Deep learning extends this concept by adding more hidden layers to the network**. Each layer can learn to represent the data at a different level of abstraction. The first hidden layer might learn to recognize simple shapes or patterns in the data, while subsequent layers might learn to combine these simple patterns into more complex ones. This hierarchy of learned features allows deep neural networks to tackle much more complex problems than their shallow counterparts.

**Key Concepts in Deep Learning**

-   *Activation Functions*: Non-linear activation functions like ReLU (Rectified Linear Unit), Sigmoid, and Tanh help deep neural networks learn complex patterns by adding non-linearity to the model, enabling it to capture a wide variety of relationships in the data.
-   *Regularization*: Techniques like dropout, L2 regularization, and batch normalization help prevent overfitting, ensuring that the model generalizes well to unseen data.
-   *Optimization Algorithms*: These are methods used to minimize the loss function, guiding how the network weights are updated during training. Examples include SGD (Stochastic Gradient Descent), Adam, and RMSprop.
-   *Backpropagation*: It's basically how gradients are calculated and weights are updated in a neural network. Starting from the output layer, the method computes the gradient of the loss function with respect to each weight by chain rule, moving backward through the network's layers. This tells us how each weight contributed to the error.

R offers several packages for deep learning, such as `keras` and `tensorflow`, which provide an interface to TensorFlow's deep learning capabilities. These packages allow for the design, training, and deployment of deep learning models with R, making it accessible for statisticians and data scientists familiar with the language.

Simply, a Deep Neural Network (DNN), or Deep Learning, is an artificial neural network that has two or more hidden layers. Even greater flexibility is achieved via composition of activation functions:

$$
y \approx \alpha+\sum_{m=1}^M \beta_m f(\alpha_m^{(1)}+\underbrace{\sum_{p=1}^P f\left(\alpha_p^{(2)}+\mathbf{X} \delta_p^{(2)}\right)}_{\text {it replaces } \mathbf{X}} \delta_m^{(1)})
$$

where $f$ is the activation function, $\alpha$ and $\beta$ are the weights, and $\delta$ are the biases. The model is trained by minimizing a loss function, such as mean squared error, using optimization algorithms like stochastic gradient descent.

# Single layer with `neuralnet` package

Single layer with different number of neurons:

```{r, warning=FALSE, message=FALSE}
library(AER)
data("CPS1985")

df <- CPS1985

# Scaling and Dummy coding
df[,sapply(df, is.numeric)] <- scale((df[, sapply(df, is.numeric)]))
ddf <- model.matrix(~.-1, data = df, contrasts.arg =
                       lapply(df[ ,sapply(df, is.factor)],
                              contrasts, contrasts = FALSE))

ddf <- as.data.frame(ddf)

# formula to pass on ANN and lm()  
w.ind <- which(colnames(ddf) == "wage")
frmnn <- as.formula(paste("wage~", paste(colnames(ddf[-w.ind]), collapse = '+')))
frmln <- as.formula(paste("wage~", "I(experience^2)", "+",
                          paste(colnames(ddf[-w.ind]), collapse = "+")))
```

```{r, warning=FALSE, message=FALSE, cache=TRUE}
library(neuralnet)

mse.test <- matrix(0, 10, 5)

for (i in 1:10) { 
  
  set.seed(i + 1)
  trainid <- sample(nrow(ddf), nrow(ddf), replace = TRUE)
  train <- ddf[trainid,]
  test <- ddf[-trainid,]
  
  # Models 
  fit.lm <- lm(frmln, data = train)
  fit.nn <- neuralnet(frmnn, data = train, hidden = 1, threshold = 0.05)
  fit.nn2 <- neuralnet(frmnn, data = train, hidden = 2, threshold = 0.05)
  fit.nn3 <- neuralnet(frmnn, data = train, hidden = 3, threshold = 0.05)
  fit.nn4 <- neuralnet(frmnn, data = train, hidden = 3, threshold = 0.05,
                       act.fct = "tanh")
 
  # Prediction errors
  mse.test[i,1] <- mean((test$wage - predict(fit.lm, test))^2)
  mse.test[i,2] <- mean((test$wage - predict(fit.nn, test))^2)
  mse.test[i,3] <- mean((test$wage - predict(fit.nn2, test))^2)
  mse.test[i,4] <- mean((test$wage - predict(fit.nn3, test))^2)
  mse.test[i,5] <- mean((test$wage - predict(fit.nn4, test))^2)
}

colMeans(mse.test)
```

# Deep Learning with `keras` package

## Deep Learning Frameworks

To train deep neural networks, we need specialized software frameworks. Here are the main ones:

| Framework | Developer | Language | Key Feature |
|-----------|-----------|----------|-------------|
| **TensorFlow** | Google | Python, C++ | The backend engine that handles tensor computation and automatic differentiation. It does the heavy math. |
| **Keras** | François Chollet (Google) | Python (R via `keras` package) | A **high-level API** that sits on top of TensorFlow. Instead of managing tensors directly, you write `layer_dense(units = 64)`. Now built into TensorFlow as `tf.keras`. |
| **PyTorch** | Meta (Facebook) | Python | TensorFlow's main competitor. More Pythonic and flexible — preferred in research. No native R support. |
| **JAX** | Google | Python | Newer framework for high-performance numerical computing with automatic differentiation. Growing in research. |

In this course, we use **Keras via TensorFlow** because it has excellent R support through the `keras` and `tensorflow` R packages.

## Why GPUs Matter

Deep learning involves **massive matrix multiplications** — multiplying weight matrices by input matrices, layer after layer, thousands of times. CPUs process these operations largely sequentially. GPUs, by contrast, have **thousands of small cores** designed for parallel computation, delivering 10–100× speedups for matrix operations.

| Platform | GPU Support | Notes |
|----------|-------------|-------|
| **Linux + NVIDIA GPU** | Best support | Native CUDA/cuDNN — the gold standard for DL training |
| **Mac with Apple Silicon (M1–M4)** | Good | Uses Metal Performance Shaders (MPS) — setup instructions below |
| **Windows + NVIDIA GPU** | Requires WSL2 | TensorFlow **dropped native Windows GPU support** in v2.11 (2022). You must install WSL2 (Windows Subsystem for Linux) to use GPU acceleration, which adds significant setup complexity. |
| **Any machine without GPU** | CPU only | Works fine for small models (like our examples), just slower for large-scale training |

For our course examples, CPU is sufficient. But for real-world deep learning (large datasets, image/text models), GPU acceleration is essential.

## Setting Up Keras on Mac (Apple Silicon)

To use Keras with GPU acceleration on a MacBook with M-series processors, you need TensorFlow configured with Apple's Metal Performance Shaders (MPS):

2.  Install TensorFlow with MPS support:

``` bash
pip install tensorflow-macos
pip install tensorflow-metal
```

3.  For R users specifically, you'll need to set up the proper environment:

    -   Install R and RStudio
    -   Install the `keras` and `tensorflow` R packages

    ``` r
    install.packages("keras")
    install.packages("tensorflow")
    ```

4.  Initialize and configure Keras to use the MPS backend:

```r
library(keras)
library(tensorflow)

# Install TensorFlow for R with MPS support
install_tensorflow(extra_packages = c("tensorflow-metal"))

# Configure to use Metal
use_backend("tensorflow")
```

5.  Verify GPU availability:

``` r
library(tensorflow)
tf$config$list_physical_devices('GPU')
```

This should show your Apple GPU device if everything is set up correctly.

When building Keras models, your code will automatically utilize the GPU acceleration provided by the Metal framework when available.

Keep in mind that while the Metal GPU acceleration provides good performance, some TensorFlow operations may not be fully optimized for Apple Silicon yet. The M chips should provide excellent performance for most deep learning tasks, especially with the properly configured environment.

The code below trains a deep neural network with two hidden layers, each containing 64 neurons. The model is trained on the training data and evaluated on the test data using the Mean Squared Error (MSE) and Mean Absolute Error (MAE) as metrics. The results show the test loss and test MAE for the model.  More details can be found here [TensorFlow for R](https://tensorflow.rstudio.com/guides/).

```{r}
library(keras3)
library(tensorflow)
library(AER)

# Check for GPU availability (if any)
print(tf$config$list_physical_devices("GPU"))
# Clear any previous Keras session
clear_session()

# Load the dataset from the AER package
data("CPS1985")
df <- CPS1985

# Scale numeric features
numeric_vars <- sapply(df, is.numeric)
df[, numeric_vars] <- scale(df[, numeric_vars])

# Create dummy variables for factor variables using model.matrix
ddf <- model.matrix(~ . - 1, data = df,
                    contrasts.arg = lapply(df[, sapply(df, is.factor)], contrasts, contrasts = FALSE))
ddf <- as.data.frame(ddf)

# Split the data into training (80%) and testing (20%) sets
set.seed(123)  # for reproducibility
train_indices <- sample(1:nrow(ddf), size = round(0.8 * nrow(ddf)), replace = FALSE)
train <- ddf[train_indices, ]
test <- ddf[-train_indices, ]

# Prepare predictor matrices and target vectors. Here 'wage' is the target.
train_x <- as.matrix(train[, -which(colnames(train) == "wage")])
train_y <- matrix(train$wage, ncol = 1)
test_x <- as.matrix(test[, -which(colnames(test) == "wage")])
test_y <- matrix(test$wage, ncol = 1)

# Define the input layer using the number of predictors
inputs <- layer_input(shape = ncol(train_x))

# Build the model using the functional API
outputs <- inputs %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1)  # Output layer for regression

# Create the model by specifying inputs and outputs
model <- keras_model(inputs = inputs, outputs = outputs)

# Compile the model with MSE loss and MAE as metric, using the RMSprop optimizer
model$compile(
  loss = "mean_squared_error",
  optimizer = optimizer_rmsprop(),
  metrics = list("mean_absolute_error")
)

# Fit the model. Setting verbose = 0 will suppress progress output.
history <- model$fit(
  x = train_x, 
  y = train_y,
  epochs = as.integer(100),
  batch_size = as.integer(128),
  validation_split = 0.2,
  verbose = 0  # change to 1 if you wish to see training progress
)

# Evaluate the model on the test set
score <- model$evaluate(
  x = test_x, 
  y = test_y,
  verbose = 1
)

cat("Test loss:", score[[1]], "\n")
cat("Test mean absolute error:", score[[2]], "\n")
```
**Test Loss**: This is the value of the loss function for our test dataset. In a regression problem like yours, where you might be using Mean Squared Error (MSE) as the loss function, this represents the average of the squares of the differences between the predicted and actual values across all observations in the test set. It's a measure of how well the model is able to predict the target variable on unseen data. A lower test loss indicates better model performance.

**Test MAE (Mean Absolute Error)**: MAE is another metric used to evaluate the performance of a regression model. It's calculated as the average of the absolute differences between the predicted and actual values across all observations in the test set. Unlike the MSE, the MAE gives a linear score, meaning all individual differences are weighted equally in the average. It's particularly useful because it's easy to interpret; for instance, an MAE of 5 would mean that, on average, the predicted values differ from the actual values by 5 units of the target variable.

```{r}
# Check the history object to see the training progress and performance metrics:
# history$history

# Convert the training history to a data frame:
history_df <- as.data.frame(history$history)

# Plot training MSE and MAE by epoch:
plot(history_df$loss, type = "l", col = "red", lwd = 2,
     ylim = range(c(history_df$loss, history_df$mean_absolute_error)),
     xlab = "Epoch", ylab = "Error", main = "Training MSE and MAE")
lines(history_df$mean_absolute_error, col = "blue", lwd = 2)
legend("topright", legend = c("MSE", "MAE"), col = c("red", "blue"), lty = 1, lwd = 2)

```


## Setting up a Keras Model: Step by Step

The code above follows a **4-step workflow** that every Keras model uses. Let's map each step to the specific code:

### Step 1: Define the Model Architecture

```r
inputs <- layer_input(shape = ncol(train_x))       # Input: 17 features

outputs <- inputs %>%
  layer_dense(units = 64, activation = "relu") %>%  # Hidden layer 1: 64 neurons, ReLU
  layer_dense(units = 64, activation = "relu") %>%  # Hidden layer 2: 64 neurons, ReLU
  layer_dense(units = 1)                            # Output: 1 neuron (regression)

model <- keras_model(inputs = inputs, outputs = outputs)
```

This uses the **functional API**: you define an input, chain layers with `%>%`, and create the model by connecting inputs to outputs. The `shape = ncol(train_x)` tells Keras how many features to expect. Each `layer_dense()` is a fully connected layer — the same dense layers we saw with `neuralnet`, but now we can stack many of them.

### Step 2: Compile the Model

```r
model$compile(
  loss = "mean_squared_error",
  optimizer = optimizer_rmsprop(),
  metrics = list("mean_absolute_error")
)
```

Compilation configures **how** the model learns:

- **Loss** = MSE — the objective to minimize (same as in `neuralnet`)
- **Optimizer** = RMSprop — the gradient descent variant (see `GradientD.qmd` for the concepts)
- **Metrics** = MAE — tracked during training for monitoring, but not used for optimization

### Step 3: Train the Model

```r
history <- model$fit(
  x = train_x, y = train_y,
  epochs = as.integer(100),
  batch_size = as.integer(128),
  validation_split = 0.2,
  verbose = 0
)
```

- **Epochs = 100**: 100 complete passes through the training data
- **Batch size = 128**: Each gradient update uses 128 observations (mini-batch gradient descent — see `GradientD.qmd`)
- **Validation split = 0.2**: 20% of training data is held out to monitor overfitting after each epoch

### Step 4: Evaluate on Test Data

```r
score <- model$evaluate(x = test_x, y = test_y)
```

Returns the test loss (MSE) and test MAE — the model's performance on data it has never seen during training.

## Keras with Python

Lets do the same thing with Python. First, we need to save the `ddf` dataframe to a CSV file so that we can load it in Python.

```{r}
write.csv(ddf, "ddf.csv", row.names = FALSE)
```

Now, we can run the following Python code to train a deep neural network on the same data and evaluate its performance on the test set.

```{python}
from tensorflow.keras import backend as K
from tensorflow.keras.models import Sequential
from tensorflow.keras.layers import Dense
from tensorflow.keras.optimizers import RMSprop
import numpy as np
import pandas as pd

# Clearing the session
K.clear_session()

# Loading the ddf dataframe from the CSV file
ddf = pd.read_csv("ddf.csv")

# Let's split the data into training and testing sets
np.random.seed(42)  # For reproducibility
sample_indices = np.random.choice(ddf.index, size = len(ddf), replace = True)
train = ddf.loc[sample_indices]
out_of_bag_indices = np.setdiff1d(ddf.index, sample_indices, assume_unique = False)
test = ddf.loc[out_of_bag_indices]

# Preparing the data
train_x = train.drop(columns=['wage']).values
train_y = train['wage'].values
test_x = test.drop(columns=['wage']).values
test_y = test['wage'].values

# Define the model
model = Sequential([
    Dense(64, activation='relu', input_shape=(train_x.shape[1],)),
    Dense(64, activation='relu'),
    Dense(1)
])

# Compile the model
model.compile(optimizer = RMSprop(), loss = 'mean_squared_error', metrics = ['mean_absolute_error'])

# Fit the model
history = model.fit(
    train_x, 
    train_y, 
    epochs=100, 
    batch_size=128, 
    validation_split=0.2, 
    verbose=0
)

# Evaluate the model
score = model.evaluate(test_x, test_y, verbose=0)

print('Test loss:', score[0])
print('Test mean absolute error:', score[1])
```

## Tuning a Keras Model

Tuning a Keras model involves experimenting with various configurations of the model's architecture, the compilation settings, and the training process to find the best performance. Here are some common ways to tune a simple Keras model:

### Adjusting Layer Configurations:

-   Number of Units: Vary the number of neurons in the Dense layers. More neurons can capture more complex patterns but can also lead to overfitting.
-   Activation Functions: Try different activation functions (e.g., relu, tanh, sigmoid) to see how they affect model performance.

### Experimenting with Optimizers:

-   Different optimizers (e.g., adam, sgd, rmsprop) have different optimization algorithms and hyperparameters (like learning rate). Switching optimizers or adjusting their parameters (e.g., learning rate) can significantly impact model performance.
-   Varying the Loss Functions: Depending on the problem type (regression vs. classification), different loss functions might be more appropriate. Even within a problem type, experimenting with different loss functions can yield better results.

### Regularization Techniques:

To combat overfitting, try adding regularization methods such as L1 or L2 regularization, dropout layers, or batch normalization. These techniques can help the model generalize better to unseen data. L1 and L2 regularization add a penalty term to the loss function based on the weights, while dropout randomly sets a fraction of input units to zero during training to prevent overfitting. Dropout is a regularization technique that helps prevent overfitting by randomly setting a fraction of input units to zero during training. This forces the model to learn more robust features and reduces the risk of overfitting.

### Model Training Adjustments:

-   Epochs and Batch Size: The number of epochs and the batch size can affect how well your model learns. Too few epochs might underfit, while too many can overfit. The batch size can influence the stability and speed of the learning process. Experiment with different values to find the optimal combination.
-   Validation Split: Adjusting the proportion of data used for validation can provide a better estimate of model performance on unseen data. A common choice is 20% of the training data.
-   Early Stopping: Implement early stopping to halt training when the validation performance stops improving. This prevents overfitting and saves time by avoiding unnecessary epochs.

### Hyperparameter Tuning Tools:

Consider using tools like Keras Tuner, Hyperopt, or scikit-learn's `GridSearchCV` or `RandomizedSearchCV` for systematic hyperparameter tuning. These tools can help you search through a large hyperparameter space efficiently and find the best configuration for your model.

In the following example, L1 and L2 regularization, dropout, and batch normalization are added to combat overfitting. An early stopping callback halts training when the validation loss hasn't improved for a specified number of epochs (patience), preventing overfitting and saving computational resources.

```{python}
from tensorflow.keras.layers import Input, Dense, Dropout, BatchNormalization
from tensorflow.keras.models import Sequential
from tensorflow.keras.optimizers.legacy import RMSprop
from tensorflow.keras.regularizers import l1_l2
from tensorflow.keras.callbacks import EarlyStopping

model = Sequential([
    Input(shape=(train_x.shape[1],)),
    Dense(64, activation='relu', kernel_regularizer=l1_l2(l1=0.01, l2=0.01)),
    Dropout(0.5),
    BatchNormalization(),
    Dense(64, activation='relu'),
    Dense(1)
])

model.compile(optimizer='adam', loss='mean_squared_error', metrics=['mean_absolute_error'])

early_stopping = EarlyStopping(monitor='val_loss', patience=10)

history = model.fit(train_x, train_y, epochs=100, batch_size=128, validation_split=0.2, callbacks=[early_stopping], verbose=0)

```

In this example, L1 and L2 regularization are added to the first hidden layer to penalize large weights, dropout is used to prevent overfitting, and batch normalization is applied to stabilize training. The early stopping callback monitors the validation loss and stops training if it doesn't improve for 10 epochs. These techniques can help improve the model's generalization performance and prevent overfitting.

```{python}
score = model.evaluate(test_x, test_y, verbose=0)

print('Test loss:', score[0])
print('Test mean absolute error:', score[1])

```

## Visualizing Model Performance

```{python}
import pydot
from tensorflow.keras.utils import plot_model

plot_model(model, to_file='model_plot.png', show_shapes=True, show_layer_names=True)
```

The `plot_model` function generates a visual representation of the model architecture, showing the connections between layers and the flow of data through the network. This can be useful for understanding the model's structure and sharing it with others.

And we can plot the training history to visualize the model's performance during training:

```{python}
import matplotlib.pyplot as plt

plt.plot(history.history['loss'], label='train loss')
plt.plot(history.history['val_loss'], label='val loss')
plt.xlabel('Epochs')
plt.ylabel('Loss')
plt.legend()
plt.show()
```

# Grid Search with Keras Tuner

**Step 1: Import Libraries** First, you import necessary libraries and modules. This includes Keras Tuner itself, TensorFlow's Keras API for building neural network models, and specific classes for model layers and the sequential model type.

**Step 2: Define the Model-Building Function** The `build_model` function is where you define the architecture of your neural network. This function takes a `HyperParameters` object (`hp`) as its argument, which is used to define the space of hyperparameters you want to explore.

-   Sequential Model: You start by initializing a Sequential model, indicating you're building a model layer by layer in sequence.
-   Dense Layers: You then add Dense layers, which are fully connected neural network layers. The number of units (neurons) in these layers is what you're tuning, so instead of hardcoding a number, you use `hp.Int('units', min_value=32, max_value=512, step=32)` to define a range of values to explore during the tuning process.
-   Compilation: Finally, you compile the model, specifying the optimizer, loss function, and metrics to monitor.

**Step 3: Initialize the Tuner**\
You instantiate a `RandomSearch` tuner object. This object will manage the tuning process, including:

-   Running trials (each trial tests a unique set of hyperparameters).
-   Evaluating model performance for each set of hyperparameters.
-   Selecting the next set of hyperparameters to test.

Parameters for the tuner include:

-   The model-building function (`build_model`).
-   The optimization objective (`val_loss`).
-   The maximum number of trials (`max_trials`) and executions per trial (`executions_per_trial`).
-   Directories for storing tuning results (`directory` and `project_name`).

**Step 4: Perform the Search** Using the `tuner.search()` method, you start the hyperparameter search process. This method requires your training data (`train_x`, `train_y`), the number of epochs for training each model, and the validation split. The tuner will train different models using varying hyperparameters within the defined space and evaluate their performance.

**Step 5: Get the Best Model and Hyperparameters** After the search concludes, you retrieve the best-performing model and its hyperparameters using `tuner.get_best_models()` and `tuner.get_best_hyperparameters()`.

```{python}
import keras_tuner as kt
from keras_tuner import RandomSearch
from keras_tuner.engine.hyperparameters import HyperParameters
from tensorflow.keras.models import Sequential
from tensorflow.keras.layers import Dense

def build_model(hp):
    model = Sequential()
    # Define the hyperparameter space for 'units'
    units = hp.Int('units', min_value=32, max_value=512, step=32)
    
    model.add(Dense(units=units, activation='relu', input_shape=(train_x.shape[1],)))
    model.add(Dense(units=units, activation='relu'))  # Reuse the 'units' variable
    model.add(Dense(1))
    model.compile(optimizer='adam', loss='mean_squared_error', metrics=['mean_absolute_error'])
    return model

tuner = RandomSearch(
    build_model,
    objective='val_loss',
    max_trials=5,
    executions_per_trial=3,
    directory='my_dir',
    project_name='helloworld'
)

tuner.search(train_x, train_y, epochs=100, validation_split=0.2, verbose=0)

best_model = tuner.get_best_models(num_models=1)[0]
best_hyperparameters = tuner.get_best_hyperparameters(num_trials=1)[0]

best_model.summary()

score = best_model.evaluate(test_x, test_y, verbose=0)

print('Test loss:', score[0])
print('Test mean absolute error:', score[1])
```

# Grid Search with R

Hyperparameter tuning in R, especially for Keras models, can be approached in several ways, but it does not directly support `keras_tuner` like Python. However, you can use a combination of custom loops or grid search strategies to achieve a similar outcome. Below is an R version that approximates the hyperparameter tuning process described in your Python script. This example uses a manual grid search approach:

```{r, warning=FALSE, message=FALSE}
library(keras3)
library(dplyr)

# Clear the Keras session
clear_session()

train_x <- as.matrix(ddf %>% select(-wage))
train_y <- matrix(ddf$wage, ncol = 1)

# Define the model-building function accepting hyperparameters
build_model <- function(units) {
  # Clear session for each model
  clear_session()
  
  # Define model using functional API instead of sequential
  inputs <- layer_input(shape = ncol(train_x))
  outputs <- inputs %>%
    layer_dense(units = units, activation = 'relu') %>%
    layer_dense(units = units, activation = 'relu') %>%
    layer_dense(units = 1)
  
  # Create model
  model <- keras_model(inputs = inputs, outputs = outputs)
  
  # Compile model
  model$compile(
    optimizer = 'adam',
    loss = 'mean_squared_error',
    metrics = list('mean_absolute_error')
  )
  
  return(model)
}

# Manual hyperparameter tuning setup
units_grid <- seq(32, 512, by = 32)
val_losses <- numeric(length(units_grid))
names(val_losses) <- as.character(units_grid)

# Grid search loop
for (units in units_grid) {
  cat("Training with units =", units, "\n")
  model <- build_model(units)
  
  history <- model$fit(
    x = train_x, 
    y = train_y,
    epochs = as.integer(100), 
    batch_size = as.integer(32),
    validation_split = 0.2,
    verbose = 0
  )
  
  val_losses[as.character(units)] <- min(history$history$val_loss)
}

# Find the best performing model
best_units <- as.numeric(names(which.min(val_losses)))
cat("Best units:", best_units, "\n")

# Build and evaluate the best model
best_model <- build_model(best_units)
best_model$fit(
  x = train_x, 
  y = train_y, 
  epochs = as.integer(100), 
  batch_size = as.integer(32),
  verbose = 0
)

# Evaluate on test data
score <- best_model$evaluate(
  x = test_x, 
  y = test_y, 
  verbose = 0
)

cat('Test loss:', score[[1]], '\n')
cat('Test mean absolute error:', score[[2]], '\n')
```

In this R script, we define a `build_model` function that creates a Keras model with a specified number of units in the Dense layers. We then set up a manual grid search loop to iterate over different values of `units` and train models with those hyperparameters. The best-performing model is selected based on the validation loss, and its performance is evaluated on the test set.
