---
title: "<span style='color: red; font-size: 50px; font-weight: bold;'>Classification with Deep Learning</span>"
subtitle: "Binary and Multi-Class Classification with Neural Networks"
author: "Dr. Aydede"
date: today
format: 
  html:
    embed-resources: true
    code-background: true
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    fig-width: 9
    fig-height: 6
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(reticulate)
use_virtualenv("/Users/yigitaydede/.virtualenvs/r-tensorflow", required = TRUE)
Sys.unsetenv("TF_USE_LEGACY_KERAS")
```

Applying deep learning to classification problems introduces some differences in model architecture, training process, and evaluation compared to regression problems. Here's an overview of these differences and some key considerations:

# Model Outcome and Activation Function

Regression models typically have a single output unit without an activation function (or with a linear activation function), predicting a continuous quantity. For binary classification, there's usually a single output unit with a sigmoid activation function. The sigmoid function maps the output to a probability score between 0 and 1, indicating the likelihood of the input belonging to the positive class.

For multi-class classification, the output layer has one unit per class and uses the softmax activation function. The softmax function converts the model outputs to probabilities that sum to 1, with each value representing the probability that the input belongs to one of the classes.

## Loss Function  
Regression models commonly use mean squared error (MSE) or mean absolute error (MAE) as the loss function. For binary classification, the binary cross-entropy loss function is standard.
For multi-class classification, categorical cross-entropy (or softmax cross-entropy) loss is typically used. This loss function compares the predicted class probabilities with the true class labels and penalizes the model based on the difference between them.

More detailed explanation of the loss functions is provided in Chapter 5.2.2. in the book. 


Cross-entropy loss is a measure of the difference between two probability distributions: the predicted distribution (output of the model) and the true distribution (one-hot encoded labels). The goal is to minimize this difference during training.  The cross-entropy loss is defined as:

$$
\text{Cross-Entropy Loss} = -\sum_{i} y_i \log(p_i)
$$

where $y_i$ is the true class label (0 or 1 for binary classification, one-hot encoded for multi-class classification), and $p_i$ is the predicted probability for class $i$.  For a binary classification, we have two possible outcomes: the true label $y$ is either 0 or 1 , and the predicted probability (from the neural network) is $p$. The binary cross-entropy loss for a single observation is:

$$
L(y, p)=-[y \log (p)+(1-y) \log (1-p)]
$$
  

**Numerical Example:** Let's compute the loss for several scenarios to see how cross-entropy penalizes predictions:

| True Label ($y$) | Predicted ($p$) | Loss $L(y,p)$ | Interpretation |
|:-:|:-:|:-:|:--|
| 1 | 0.9 | $-\log(0.9) = 0.105$ | Confident & correct → low loss |
| 1 | 0.5 | $-\log(0.5) = 0.693$ | Uncertain → moderate loss |
| 1 | 0.1 | $-\log(0.1) = 2.303$ | Confident & wrong → high loss |
| 0 | 0.1 | $-\log(0.9) = 0.105$ | Confident & correct → low loss |
| 0 | 0.9 | $-\log(0.1) = 2.303$ | Confident & wrong → high loss |

Notice: when the model is confident and correct ($p=0.9$ for $y=1$), the loss is only 0.105. But when it's confident and wrong ($p=0.1$ for $y=1$), the loss jumps to 2.303 — a 22× penalty. This steep penalty for confident wrong predictions is what drives the model to produce well-calibrated probabilities.

- If the true label $y=1$, the loss becomes $L(1, p)=-\log (p)$. As $p$ (the predicted probability of the positive class) decreases, $-\log (p)$ increases, meaning the loss increases when the model's prediction is further from the actual class. 
- If the true label $y=0$, the loss becomes $L(0, p)=-\log (1-p)$. As $p$ increases, $-\log (1-$ $p$ ) increases, meaning the loss increases when the model's prediction is incorrectly confident in the positive class.

This loss function effectively penalizes predictions that are confident and wrong, encouraging the model to produce probabilities close to the actual labels.  And the total loss is

$$
J=\sum_{i=1}^n y_i \log p\left(x_i\right)+\left(1-y_i\right) \log \left(1-p\left(x_i\right)\right),
$$
where for each observation,
  
$$
p\left(\mathbf{x}_{\mathbf{i}}\right)=\frac{e^{\beta_0+\beta x_i}}{1+e^{\beta_0+\beta x_i}}
$$
If the activation function is the sigmoid function.

## Probabilistic Model and Thresholding
Neural networks can be considered probabilistic in the context of classification. The output layer (using sigmoid or softmax) provides probabilities indicating the likelihood of each class. Although the raw output of the network provides these probabilities, making a final class decision often requires thresholding

As before, we cannot use a common threshold, like 0.5, for the sigmoid output. We need to rely on AUC. In multi-class classification, the class with the highest probability (from the softmax output) is typically selected as the model's prediction.

Unlike regression, where you might use MSE or MAE, classification models are often evaluated using accuracy, precision, recall, F1 score.  We use, however, a complex metrics like ROC-AUC or confusion matrices, depending on the specific requirements of the problem.

# Classification with `neuralnet` Package

## Binary Classification. 
The goal is to predict whether a client will subscribe to a term deposit (variable $y$) based on various features such as age, job, marital status, education, and economic indicators.

```{r, warning=FALSE, message=FALSE}
library(caret)
library(dplyr)
library(readr)
library(skimr)
library(ROCR)

df <- read_delim("bank.csv", delim = ";",
                   escape_double = FALSE, trim_ws = TRUE)

# Scaling and Dummy coding
df[,sapply(df, is.numeric)] <- scale((df[, sapply(df, is.numeric)]))
ddf <- model.matrix(~.-1, data = df, contrasts.arg =
                       lapply(df[ ,sapply(df, is.factor)],
                              contrasts, contrasts = FALSE))

# Colnames have to be cleaned
ddf <- as.data.frame(ddf)
clean_names <- make.names(names(ddf), unique = TRUE)
names(ddf) <- clean_names

# formula to pass on ANN and lm()  
y.ind <- which(colnames(ddf) == "yyes")
frm <- as.formula(paste("yyes~", paste(colnames(ddf[-y.ind]), collapse = '+')))
```

```{r, warning=FALSE, message=FALSE, cache=TRUE}
library(neuralnet)

set.seed(1)
trainid <- sample(nrow(ddf), nrow(ddf), replace = TRUE)
train <- ddf[trainid, ]
test <- ddf[-trainid, ]

fit.ln <- glm(frm, data = train, family = binomial(link = "logit"))  
fit.nn <- neuralnet(frm,
                    data = train,
                    hidden = c(4),
                    threshold = 0.05,
                    linear.output = FALSE,
                    err.fct = "ce")
#Predictions
phat.ln <- predict(fit.ln, test, type = "response")
phat.nn <- predict(fit.nn, test, type = "response")

#AUC for Logisitic Regression
pred_rocr1 <- ROCR::prediction(phat.ln, test$yyes)
auc_ROCR1 <- ROCR::performance(pred_rocr1, measure = "auc")
auc_ROCR1@y.values[[1]]

#AUC for Neural Network
pred_rocr2 <- ROCR::prediction(phat.nn, test$yyes)
auc_ROCR2 <- ROCR::performance(pred_rocr2, measure = "auc")
auc_ROCR2@y.values[[1]]
```

## Multi-Class Classification

This dataset is public available for research. The details are described in (P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.)

```{r, warning=FALSE, message=FALSE}
dfr <- read.csv("wineQualityReds.csv", header = TRUE)
dfr <- dfr[, -1] # removing the index
table(dfr$quality)
```

```{r}
# Let's remove the outlier qualities:
indo <- which(dfr$quality == "3" | dfr$quality == "8")
dfr <- dfr[-indo, ]
dfr$quality <- as.factor(dfr$quality)
table(dfr$quality)
```

```{r}
# Scaling and Dummy coding
dfr[, sapply(dfr, is.numeric)] <-
  scale((dfr[, sapply(dfr, is.numeric)]))

ddf <- model.matrix( ~ quality - 1, data = dfr)
y.ind <- which(colnames(dfr) == "quality")
dfr <- dfr[, -y.ind] # removing 'quality`
df <- cbind(ddf, dfr)

frm <- as.formula(paste(
  paste(colnames(ddf), collapse = '+'),
  "~",
  paste(colnames(dfr), collapse = '+')
))

frm
```


```{r}
library(neuralnet)

ind <- sample(nrow(df), nrow(df) * 0.7)
train <- df[ind, ]
test <- df[-ind, ]

fit.nn <-
  neuralnet(
    frm,
    data = train,
    hidden = c(3, 2),
    threshold = 0.05,
    linear.output = FALSE,
    err.fct = "ce"
  )

plot(fit.nn, rep = "best")
```

```{r}
phat <- predict(fit.nn, test)
head(phat)
```


```{r}
# Assigning label by selecting the highest phat
label.hat <- t(apply(phat, 1, function(x) as.numeric(x == max(x))))
head(label.hat)
```

  
```{r}
# Confusion Table
pred <- apply(phat, 1, which.max)
fck <- colnames(test)[1:4]
predicted <- fck[pred]

act <- apply(test[ ,1:4], 1, which.max)
actual <- fck[act]

table(predicted, actual)
```

Now AUC

```{r, warning=FALSE, message=FALSE}
library(pROC)

test_labels <- test[ ,1:4]  # Extracting the first 4 columns as the true labels

# Calculate AUC for each class and store it
aucs <- sapply(1:ncol(test_labels), function(i) {
  roc_auc <- roc(test_labels[ ,i], phat[ ,i])$auc
  return(roc_auc)
})

# Calculate the average AUC across all classes
unlist(aucs)
mean_auc <- mean(aucs)

print(mean_auc)
```

# `keras` Package

First the preprocessing

```{r}
library(keras3)
library(tensorflow)
library(caret) # for dummyVars
library(dplyr) # for data manipulation

# Clearing the workspace
rm(list = ls())

# Clearing the Keras session
clear_session()

# Load the data 
dfr <- read.csv("wineQualityReds.csv", header = TRUE)
dfr <- dfr[, -1] # removing the index column
indo <- which(dfr$quality == 3 | dfr$quality == 8) # Ensure quality is treated as numeric, not string
dfr <- dfr[-indo, ]

# Convert quality to a factor for one-hot encoding
dfr$quality <- as.factor(dfr$quality)

# One-hot encode the quality variable
quality_one_hot <- model.matrix(~ quality - 1, data = dfr)

# Prepare the data excluding the original quality column
df_transformed <- dfr[, -which(names(dfr) == "quality")]

# Scale the numeric features (important for neural networks!)
df_transformed <- as.data.frame(scale(df_transformed))

# Splitting the dataset into training and test sets
set.seed(123) # For reproducibility
indexes <- sample(1:nrow(dfr), size = 0.8 * nrow(dfr), replace = FALSE) # Randomly sample 80% of the data

train_x <- df_transformed[indexes, ]
test_x <- df_transformed[-indexes, ]
train_y <- quality_one_hot[indexes, ]
test_y <- quality_one_hot[-indexes, ]

# Define the input layer using the number of predictors
inputs <- layer_input(shape = ncol(train_x))

# There are 4 classes after your preprocessing
outputs<- inputs %>%
  layer_dense(units = 32, activation = 'relu') %>%
  layer_dense(units = 32, activation = 'relu') %>%
  layer_dense(units = 4, activation = 'softmax') # Update this to match the number of classes

# Create the model by specifying inputs and outputs
model <- keras_model(inputs = inputs, outputs = outputs)

# Compile the model for multi-class classification
model$compile(
  loss = 'categorical_crossentropy', # Correct loss function for multi-class
  optimizer = 'adam',
  metrics = list(metric_binary_accuracy(), metric_auc(name = "auc")) # Accuracy is a common metric for classification
)

# Fit the model
history <- model$fit(
  x = as.matrix(train_x), 
  y = train_y,
  epochs = as.integer(100), 
  batch_size = as.integer(128), 
  validation_split = 0.2, 
  verbose = 0
)

# Predictions as probabilities
predictions <- model$predict(as.matrix(test_x))
head(predictions)

# Evaluate the model on the test set
test_x_matrix <- as.matrix(test_x)
score <- model$evaluate(test_x_matrix, test_y, verbose = 1)

cat("Test loss (categorical crossentropy):", score[[1]], "\n")
cat("Test binary accuracy:", score[[2]], "\n")
cat("Test AUC:", score[[3]], "\n")
```

Now AUC

```{r, warning=FALSE, message=FALSE}
library(pROC)

# Initialize an empty vector to store AUC for each class
aucs <- c()

# Calculate AUC for each class
for (i in 1:ncol(test_y)) {
  roc_res <- roc(response = test_y[, i], predictor = predictions[, i])
  aucs[i] <- auc(roc_res)
}

# Calculate mean AUC across all classes
mean_auc <- mean(aucs)

print(aucs)
print(mean_auc)

```

::: {.callout-important}
## Why Are the Two AUC Values Different?
You may notice that the **Keras AUC** (from `model$evaluate()`) is higher than the **pROC AUC** (from `mean_auc`). These are computed differently:

- **Keras `metric_auc()`** uses a batch-wise **approximation** with binned thresholds (default 200 bins), treating each softmax output as an independent binary classification. This approximation can overestimate AUC, especially with smaller test sets.
- **pROC `roc()`** computes the **exact** AUC using the full ROC curve (trapezoidal rule) for each class, then takes a simple macro-average.

The pROC calculation is the more standard and reliable metric for final evaluation. Keras AUC is primarily useful for **monitoring trends during training** (is AUC going up or down across epochs?), not as a definitive performance measure.
:::

::: {.callout-note}
## AUC as a Training Metric
In earlier versions of Keras, AUC couldn't be tracked during training because it requires the entire dataset's predictions to compute accurately (unlike accuracy, which works batch-wise). Since TensorFlow 2.2.0+, you can include AUC as a metric during compilation using `tf$keras$metrics$AUC()`.
:::

## Keras Classification with Python

Let's replicate the same multi-class classification in Python. First, we save the wine data from R:

```{r}
# Save the wine quality data for Python
dfr_save <- read.csv("wineQualityReds.csv", header = TRUE)
write.csv(dfr_save, "wineQualityReds_clean.csv", row.names = FALSE)
```

```{python}
from tensorflow.keras import backend as K
from tensorflow.keras.layers import Dense, Input
from tensorflow.keras.models import Model
import numpy as np
import pandas as pd

# Clear session
K.clear_session()

# Load and preprocess
df = pd.read_csv("wineQualityReds.csv")
df = df.iloc[:, 1:]  # Remove index column
df = df[~df['quality'].isin([3, 8])]  # Remove outlier quality levels

# Separate features and target
X = df.drop(columns=['quality']).values
y = df['quality'].values

# Scale features (manual z-score standardization)
X = (X - X.mean(axis=0)) / X.std(axis=0)

# One-hot encode the target using pandas
y_onehot = pd.get_dummies(y).values
n_classes = y_onehot.shape[1]

# Train/test split
np.random.seed(123)
indices = np.random.permutation(len(X))
split = int(0.8 * len(X))
train_x, test_x = X[indices[:split]], X[indices[split:]]
train_y, test_y = y_onehot[indices[:split]], y_onehot[indices[split:]]

# Build model (same architecture as R version)
inputs = Input(shape=(train_x.shape[1],))
x = Dense(32, activation='relu')(inputs)
x = Dense(32, activation='relu')(x)
outputs = Dense(n_classes, activation='softmax')(x)

model = Model(inputs=inputs, outputs=outputs)

# Compile with categorical crossentropy (multi-class)
model.compile(
    optimizer='adam',
    loss='categorical_crossentropy',
    metrics=['accuracy']
)

# Train
history = model.fit(
    train_x, train_y,
    epochs=100,
    batch_size=128,
    validation_split=0.2,
    verbose=0
)

# Evaluate
score = model.evaluate(test_x, test_y, verbose=0)
print(f'Test loss: {score[0]:.4f}')
print(f'Test accuracy: {score[1]:.4f}')

# Predictions
predictions = model.predict(test_x, verbose=0)
print(f'Prediction shape: {predictions.shape} (samples × {n_classes} classes)')
print(f'First prediction (probabilities): {predictions[0].round(4)}')

# Calculate AUC for each class using TensorFlow's built-in AUC metric
import tensorflow as tf

auc_per_class = []
for i in range(test_y.shape[1]):
    auc_metric = tf.keras.metrics.AUC()
    auc_metric.update_state(test_y[:, i], predictions[:, i])
    auc_per_class.append(auc_metric.result().numpy())

print(f'AUC per class: {[round(float(a), 4) for a in auc_per_class]}')
print(f'Mean AUC: {np.mean(auc_per_class):.4f}')
```

::: {.callout-note}
## Why Do R and Python AUC Values Differ?
The R and Python sections produce different AUC values despite using the same dataset and architecture. This is expected because: (1) R and Python have **different random number generators**, so `set.seed(123)` and `np.random.seed(123)` produce different train/test splits; (2) R's `scale()` uses sample SD ($n-1$ denominator) while NumPy's `std()` uses population SD ($n$ denominator); and (3) each model starts with different random weight initializations. The two sections demonstrate the **same approach** in two languages — they are not expected to produce identical numbers.
:::

The key differences from the regression model in `DL.qmd`:

- **Output layer**: `Dense(4, activation='softmax')` instead of `Dense(1)` — one neuron per class with softmax
- **Loss function**: `categorical_crossentropy` instead of `mean_squared_error`
- **Evaluation metric**: Accuracy and AUC instead of MAE

